---
title: "Iterating Over Functions"
editor: source
---

In this unit, you’ll review for iteration---repeatedly performing the same
function on different inputs. 

Iteration in R generally tends to look a bit different from other programming
languages. Much of iteration we get for free! For example, if you want to double
a numeric vector `x` in R, you can just write `2 * x`, whereas in many other
languages you would need to explicitly double each element of `x` using some
sort of `for` loop.

In R, there are generally two methods for iteration---`for()` loops and 
functionals. We will start with a review of `for()` loops before hopping over
to functionals. 

```{r}
#| label: packages
#| include: false
library(tidyverse)
```

# `for` loops

In R, `for` loops have the following general structure:

```{r for_syntax, eval=FALSE}
for (i in some_vector) {
    # Code to do stuff with i
}
```

`some_vector` can be **any** vector, including:

- An indexing vector: `1:3`
- A character vector: `c("group1", "group2", "group3")`
- A vector of any other class

```{r}
#| label: for-examples

groups <- c("group1", "group2", "group3")

for (i in 1:3) {
    print(groups[i])
}

for (g in groups) {
    print(g)
}
```

## `for()` loop Indices

The `seq_along()` function generates an integer sequence from 1 to the length of
the vector supplied. A nice feature of `seq_along()` is that it generates an
empty iteration vector if the vector you're iterating over itself has length 0.

```{r}
#| label: for-seq-along

seq_along(groups)
no_groups <- c()
seq_along(no_groups)

for (i in seq_along(groups)) {
    print(groups[i])
}

for (i in seq_along(no_groups)) {
    print(no_groups[i])
}
```

Closely related to `seq_along()` is `seq_len()`. While `seq_along(x)` generates
an integer sequence from 1 to `length(x)`, `seq_len(x)` takes `x` itself to be a
length:

```{r}
#| label: for-seq-len

seq_len(3)
seq_len(0)

for (i in seq_len(length(groups))) {
    print(groups[i])
}

for (i in seq_len(length(no_groups))) {
    print(no_groups[i])
}
```

## Dataframe Indices

`seq_len()` is useful for iterating over the rows of a data frame because
`seq_along()` would iterate over columns:

```{r}
#| label: small-data

small_data <- tibble(a = 1:2, 
                     b = 2:3, 
                     c = 4:5)
small_data
```

```{r}
#| label: seq-along-length

for (col in seq_along(small_data)) {
    print(col)
}

for (r in seq_len(nrow(small_data))) {
    print(r)
}

```

## Storing Objects

Often we'll want to store output created during a `for` loop. We can create
storage containers with the `vector()` function:

```{r}
#| label: initialize-vectors

char_storage <- vector("character", 
                       length = 3)
char_storage

num_storage <- vector("numeric", 
                      length = 3)
num_storage

list_storage <- vector("list", 
                       length = 3)
list_storage
```

Our loop can then store objects created at each iteration stage:

```{r}
#| label: store-results

for (i in seq_len(3)) {
    char_storage[i] <- str_c("Number: ", i)
    num_storage[i] <- 2*i
    list_storage[[i]] <- i # Note the [[ for subsetting here
}

char_storage
num_storage
list_storage
```



## Initial Object Size

::: stat541
:::

Notice that we initialized `char_storage`, `num_storage`, and `list_storage` to
all be the same length as our iteration vector (`seq_len(3)`). Technically, this
wasn't necessary. We could have been much lazier when we initialized these 
objects: `list_storage <- NA`. Why didn't we do this?

When you don't initialize an object as a specific size (e.g., `length = 3`) 
then at every stage of the `for()` loop R needs to append a new index to that
object. This **really** slows down your `for()` loop!

Here is a comparison of the run time between two functions that both store the 
index of the iteration. The first function (`do_stuff_allocate`) sets the 
initial object size (`vector(length = reps)`) before the loop is run. The 
second function (`do_stuff_tackon`) does not set an initial object size and 
instead initializes `results` as a vector of length 1 (`results <- NA`). 

```{r}
#| echo: true
#| label: tack-on-comparison-functions

do_stuff_allocate <- function(reps) {
  results <- vector(length = reps)
  for (i in seq_len(reps)) {
    results[i] <- i
  }
  return(results)
}

do_stuff_tackon <- function(reps) {
   results <- NA
  for (i in seq_len(reps)) {
    results <- c(results, i)
  }
  return(results)
}
```

Let's see how these two functions compare in their run time:

```{r}
#| echo: false
#| warning: false
#| label: tack-on-comparison-runtime

microbenchmark::microbenchmark(
  do_stuff_allocate(reps = 1000),
  do_stuff_tackon(reps = 1000)
  ) |> 
  as.data.frame() |> 
  group_by(expr) |> 
  summarize(min = min(time), 
            mean = mean(time), 
            median = median(time), 
            max = max(time)
            ) |> 
  mutate(
    across(.cols = min:max, 
           .fns = scales::label_comma(accuracy = 1, suffix = " µs")
           ), 
    expr = str_extract(expr, pattern = "^[^(]+")
  ) |> 
  pivot_longer(cols = min:max, 
               names_to = "stat", 
               values_to = "value") |> 
  pivot_wider(names_from = expr, 
              values_from = value) |> 
  rename(Statistic = stat) |> 
  gt::gt()

```

Lesson: **You should always initialize your object as the size you want it to be.**

::: {.callout-check-in}
1. Write a `for()`-loop that prints the even numbers from 1:20. 

2. Can you produce the same output with the `seq()` function?

```{r}
#| label: for-ex-solutions-1
#| echo: false
#| eval: false

for (i in seq_len(10)) {
    print(2*i)
}

seq(from = 2, to = 20, by = 2)
```


3. Write a `for()`-loop that iterates over the `month.name` vector (built-in to
base R) and stores a character vector of output containing strings like 
"Month 1: January", "Month 2: February". 

4. Can you produce the same output with `str_c()` only?

```{r}
#| label: for-ex-solutions-2
#| eval: false
#| echo: false

month_strings <- vector("character", length = length(month.name))

for (i in seq_along(month.name)) {
    month_strings[i] <- str_c("Month ", i, ": ", month.name[i])
}
month_strings

str_c("Month ", 1:12, ": ", month.name)
```

5. Write a `for()`-loop that store the `class()` (type) of every column in the
`mpg` data frame. 

```{r}
#| label: col_classes_soln
#| eval: false
#| echo: false

col_classes <- vector("character", 
                      ncol(mpg)
                      )

for (i in seq_along(mpg)) {
    col_classes[i] <- class(mpg[[i]])
}
```
:::

# Iteration with Functionals

> A functional is a function that takes a function as an input and returns a
> vector as output. 
> - Hadley Wickham

::: {.callout-required-reading}
[*Advanced R* Chapter 9: Functionals](https://adv-r.hadley.nz/functionals.html)
:::

`purrr` is a `tidyverse` package that provides several useful functions for
iteration. The main advantages of `purrr` include:

- Improved readability of R code
- Reduction in the "overhead" in writing a `for()`-loop (creating storage
containers and writing the `for (i in ...)`)

In `purrr`, we can use the family of `map()` functions to apply a function to
each element of a list or vector. We can think of this as mapping an input (a
list or vector) to a new output via a function. Let's look at the [`purrr` cheatsheet](https://rstudio.github.io/cheatsheets/purrr.pdf) to look at
graphical representations of how these functions work.

- `map()` returns a list
- `map_chr()` returns a character vector
- `map_lgl()` returns a logical vector
- `map_int()` returns an integer vector
- `map_dbl()` returns a numeric vector
- `map_vec()` returns a vector of a different (non-atomic) type (like dates or
factors)

## A Single Output

To get the `class()` of each data frame column, `map_chr()` is the function we
want because the class of a variable is a strings (e.g., `"logical"`).

```{r}
#| label: purrr_col_classes

map_chr(mpg, .f = class)
```

::: {.callout-note}
# List Input

The first input of `map()` functions must be a list. A dataframe is a special
type of list, where the columns are the different elements of the list 
(e.g., `mpg[["manufacturer"]]`). `map_chr()` iterates over the **columns**
(elements) of the `mpg` dataframe (list). 
:::

Let's get the class of each variable in `diamonds`:

```{r}
#| label:  purrr_col_classes_error
#| error: true

map_chr(diamonds, .f = class)
```

## Multiple Outputs

What happened!? `map_chr()` was expecting to create a character vector with 
**one** element per column in `diamonds`. But something happened in column 2
with the `cut` variable. Let's figure out what happened:

```{r}
#| label: class_diamonds_cut

class(diamonds$cut)
```

Ah! `cut` has **two** classes. In this case, `map()` (which returns a list) is
the best option because some variables have multiple classes:

```{r}
#| label: purrr_col_classes_no_error

map(diamonds, .f = class)
```

The error we encountered with `map_chr()` is a nice feature of `purrr` because
it requires us to be very sure of the type of output we are getting. Failing
loudly is vastly preferable to getting unexpected outputs silently!

## Combining with Tidy Selection

We can combine `map_*()` functions with tidy selection for some powerful
variable summaries that require much less code than `for()` loops.

```{r}
#| label:  purrr_examples_by_col

diamonds %>% 
  select(where(is.numeric)) %>% 
  map_dbl(.f = mean)

diamonds %>% 
  select(!where(is.numeric)) %>% 
  map_int(.f = n_distinct)
```

::: {.callout-check-in}
6. Using a map function reproduce this table which indicates whether a given 
column is numeric.

```{r}
#| echo: false
#| label: map-lgl-solution

diamonds |> 
  map_lgl(.f = is.numeric)
```

7. Using a map function reproduce this table which indicates how many levels
are included in each categorical variable.

```{r}
#| echo: false
#| label: map-int-solution

diamonds |> 
  select(where(is.factor)) |> 
  map_int(.f = n_distinct)
```

:::

## Multiple Inputs

`purrr` also offers the `map2()` and `pmap()` family of functions that take
**multiple** inputs and loop over them simultaneously. The [`purrr` cheatsheet](https://rstudio.github.io/cheatsheets/purrr.pdf) provides nice 
graphical representations of how these functions work.

For all the examples below, I'm going to work with this dataset of strings:

```{r}
#| label: pmap_ex1

string_data <- tibble(
    string = c("apple", "banana", "cherry"),
    pattern = c("p", "n", "h"),
    replacement = c("P", "N", "H")
)

string_data
```

### Two Inputs

The `str_detect()` function takes two arguments: a `string` and a `pattern`
to detect. The function returns logical values (`TRUE`, `FALSE`) indicating 
whether the pattern was detected in the string. Let's use this to see how the 
`map2_lgl()` function works!

```{r}
#| label: map2-example

map2_lgl(
  .x = string_data$string,
  .y = string_data$pattern,
  .f = str_detect
)
```

This should look a bit different, we are using `string_data$string` to 
input the variables we want into `map2()` rather than piping (`|>`) the
variables into the function. 

If we had tried to use the pipe operator, we would have gotten the following 
error message:

```{r}
#| error: true
#| label: bad-variable-specification

string_data |> 
  map2_lgl(
    .x = string,
    .y = pattern,
    .f = str_detect
    )
```
This error is because the pipe operator is inputting the entire `string_data`
dataframe into the first argument of `map2_lgl()` (`.x`). Looking at the 
documentation for `map2()`, the `.x` and `.y` arguments should be specified as
a pair of vectors, not dataframes. 

If we wanted to use the pipe operator, we would need to join this with a data
masking function (e.g., `mutate()`) that allows us to reference variable names
for inputs into functions:

```{r}
#| label: data-mask-map2

string_data |> 
  mutate(found = map2_lgl(
    .x = string,
    .y = pattern,
    .f = str_detect
    )
  )

```

::: stat541
:::

### Three or More Inputs

Now that we've conquered two inputs, let's try three! The `str_replace_all()`
function takes three inputs: a `string`, a `pattern` to look for, and a 
`replacement` pattern (to use when the `pattern` is found).

```{r}
#| label: pmap-name-match

string_data

pmap_chr(string_data, .f = str_replace_all)
```

Note how the column names in `string_data` exactly match the argument names in
`str_replace_all()`. The iteration that is happening is across rows, and the
multiple arguments in `str_replace_all()` are being matched by name. So, the 
first row is effectively running 
`str_replace_all(string = "apple", pattern = "p", replacement = "P")`, and
similarly for the second and third row. 

What if the column names didn't match? Well, we would need to take a similar
approach to what we did with `map2()`: 

```{r}
#| label: non-matchin-names

string_data <- string_data |> 
  rename(word = string, 
         look_for = pattern, 
         replace_with = replacement)

pmap_chr(
  .l = list(string = string_data$word,
       pattern = string_data$look_for,
       replacement = string_data$replace_with),
  .f = str_replace_all
)
```

The main difference here is there is **one** argument (`.l`) where we specify 
the inputs to the function (instead of `.x` and `.y`). This argument is a list, 
where the elements of the list should take on the same names as the function 
arguments. 

Similar to before, if I wanted to use `pmap()` to modify the `word` column of 
the dataset, I would need to pair it with `mutate()`:

```{r}
#| label: pmap-mutate

string_data |> 
  mutate(word = pmap_chr(
  .l = list(string = word,
       pattern = look_for,
       replacement = replace_with),
  .f = str_replace_all
  )
)
```



::: {.callout-check-in}
8. The function `str_c()` concatenates strings. Create a small example that uses
`map2_chr()` to combine two character vectors element by element. Each element
of the first vector should be combined with the corresponding element of the
second vector, separated by a space.

```{r}
#| label: map2-exercise-tibble
#| echo: true
#| eval: false

string_data <- tibble(
  first  = c("apple", "banana", "cherry"),
  second = c("pie", "bread", "jam")
)
```

```{r}
#| label: map2-exercise
#| echo: false
#| eval: false

map2_chr(
  string_data$first,
  string_data$second,
  function(x, y) str_c(x, y, sep = " ")
)
```

9. The function `str_sub()` takes three arguments: `string`, `start`, and `end`.
Create a small tibble containing the three inputs required for `str_sub()`. 
**Note that the names of these columns must be the same as the names of the
arguments to `str_sub()`!**

- a string,
- a starting position, and
- an ending position.

```{r}
#| label: pmap-exercise-tibble
#| echo: false
#| eval: false

string_data <- tibble(
    string = c("apple", "banana", "cherry"),
    start = c(1, 2, 4),
    end = c(2, 3, 5)
)
```

10. Now use `pmap_chr()` to apply `str_sub()` row-by-row to the tibble created
above.

```{r}
#| label: pmap-exercise-solution
#| echo: false
#| eval: false

pmap_chr(.l = string_data, .f = str_sub)
```
:::
