---
title: "JSON Data & APIs"
editor: source
---

<!-- NEED TO ADD: -->
<!-- Grabbing content from an API query with $content -->
<!-- Using these functions: -->
<!-- - rawToChar() -->
<!-- - as.tbl_json() -->
<!-- - fromJSON()  -->

<!-- Discussion of why you would want to use tools from jsonlite or tidyjson instead -->
<!-- of unnest_wider or unnest_longer -->

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: packages

library(tidyverse)
library(jsonlite)
```

# Tabular Data

It's likely that a huge portion of the datasets you've worked with in your life
thus far are very well structured and organized to the point of being
**tabular**. That is, the dataset consists of a single table with rows and
columns. If you're not familiar with the concept of **tidy data** I suggest you
check out this [`tidyr` vignette](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html)
on what it means for a dataset to be "tidy." Most of your datasets have likely
even been tidy:

-   Each variable forms a column
-   Each observation forms a row

Almost everyone's experiences beyond this probably involved taking multiple
**tidy** datasets and combining them in some way, but the overall dataset still
fit into this tabular structure. These often came in the form of TXT, CSV,
Excel, or TSV files.

Our world is rich with data and a great deal of it doesn't come in nice, neat
tables...BUT a lot of it is still extremely structured and well organized.

## Structured, but Non-Tabular Data

There is another very popular data format called "JSON". JSON is short for
**J**ava**S**cript **O**bject **N**otation and is a syntax for storing and
exchanging data, but data in a much broader sense than we're used to thinking
about.

In the middle of a statistics-related course when I hear the word "data" my mind
instinctively envisions something tabular with which we could do some sort of
visualization or analysis with. However, in the broadest sense, the word "data"
really just describes pieces of information which encapsulates everything from
movie theater schedules when you look up showtimes on a website to what your
cell phone *uses* as part of your "data plan."

While the website for your favorite movie theater may not have had your
visualization and analysis plans in mind, their data may still be of interest
to us. Many companies and organizations have built tools that allow access to
their data in a more streamlined way, and JSON has become a ubiquitous format
for many of these data sources to use.

# Hierarchical Data

One of the reasons the JSON format is so important and useful is because it can
accommodate very complex types and shapes of data. So while it may not always be
workable into a tabular form, it still maintains a high degree of structure that
we can exploit!

::: {.callout-required-reading}
[*R for Data Science*, Chapter 23](https://r4ds.hadley.nz/rectangling.html)

**Only read Sections 1-4 for now!**
:::

::: {.callout-check-in}
1. What happens when you use `unnest_wider()` with unnamed list-columns like 
`df2`? 

```{r}
#| label: df2-example
#| echo: true

df2 <- tribble(
  ~x, ~y,
  1, list(11, 12, 13),
  2, list(21),
  3, list(31, 32),
)
```

2. What argument is now necessary to correctly unnest `df2`? Specifically, the
column names should be `y_1`, `y_2`, and `y_3`? 

```{r}
#| error: true
#| label: unnest-missing-arg

unnest_wider(df2, col = y)
```

3. How does `unnest_wider()` reconcile columns with different lengths?

```{r}
#| include: false
#| label: code-for-q1
#| eval: false

unnest_wider(df2, col = y, names_sep = "_")

```
:::

# JSON Data

::: {.callout-required-reading}
[*R for Data Science*, Chapter 23](https://r4ds.hadley.nz/rectangling.html)

**Now read Sections 5 & 6!**
:::

::: {.callout-check-in}
4. JSON Objects are held in \_\_\_\_\_.

-   parentheses
-   double quotes
-   square brackets
-   curly braces

5.  While it may seem tedious compared to our usual tabular forms of data, JSON
data is easy to read because they \_\_\_\_\_.

-   come in name / value pairs
-   are separated by "-----"
-   come in table / name / value triplets

6.  Create rectangular versions of `df_col` and `df_row` (shown below). 

::: columns
::: {.column width="48%"}
```{r}
#| echo: true
#| label: code-for-json-col

json_col <- parse_json('
  {
    "x": ["a", "x", "z"],
    "y": [10, null, 3]
  }
')

df_col <- tibble(json = list(json_col)) 

df_col

```

:::

::: {.column width="4%"}
:::

::: {.column width="48%"}
```{r}
#| echo: true
#| label: code-for-json-row

json_row <- parse_json('
  [
    {"x": "a", "y": 10},
    {"x": "x", "y": null},
    {"x": "z", "y": 3}
  ]
')

df_row <- tibble(json = json_row)

df_row
```

:::
:::

**In both cases, your unnested data should look like this:**

```{r}
#| echo: false
#| label: solution-for-json-col

df_col %>% 
  unnest_wider(col = json) %>% 
  unnest_longer(col = c(x, y)) |> 
  gt::gt()
```

```{r}
#| echo: false
#| eval: false
#| label: solution-for-json-row

df_row %>% 
  unnest_wider(col = json)
```

:::

# Working with JSON Data

The chapter covers a few functions from the `jsonlite` package, but there are 
several other packages for working with JSON data in R (e.g., `rjson`, 
`rjsonio`). These packages automatically transform JSON data into nested R
lists or complex data frames. However, working with these complex objects can be
difficult.

The `tidyjson` package takes a different approach to structuring JSON data into
tidy data frames. Similar to `tidyr`, `tidyjson` builds a grammar for
manipulating JSON into a tidy table structure.

`tidyjson` is based on the following principles:

-   Work on a single JSON document, or on a collection of related documents
-   Create pipelines with the pipe operator (`%>%`), producing code that can be
read from left to right
-   Guarantee the structure of the data produced, even if the input JSON
structure changes (with the exception of `spread_all()`)
-   Work with arbitrarily nested arrays or objects
-   Handle 'ragged' arrays and / or objects (varying lengths by document)
-   Ensure edge cases are handled correctly (especially empty data)
-   Integrate seamlessly with `dplyr`, allowing `tbl_json` objects to pipe in
and out of `dplyr` verbs where reasonable

Let's use the `companies` JSON dataset from the **tidyjson** package to explore
tools from the **tidyjson** package. This exploration parallels the `worldbank`
examples from the [**tidyjson** vignette](https://cran.r-project.org/web/packages/tidyjson/vignettes/introduction-to-tidyjson.html).

```{r}
#| label: tidy-json-companies-data
#| message: false
#| warning: false

library(tidyjson)

# Companies is a long character vector
companies %>% 
  str()
```

### Step 1 -- Use `spread_all()` to spread the values of a JSON object

```{r}
#| label: explore-worldbank-data
#| message: false

companies_tidy <- spread_all(companies)
```

::: {.callout-check-in}
8. How does `spread_all()` know how many columns to make? 

9. Is the JSON data retained after you have spread the data?
:::

### Step 2 -- Find the values that were not able to be spread

To identify the names of an object, their type, and frequency we use the
following pipeline:

```{r}
#| eval: false
#| echo: true
#| label: process-for-finding-missed-values

name_of_data %>% 
  gather_object() %>% 
  json_types() %>% 
  count(name, type)
```

The `gather_object()` function collapses a JSON object into name-value pairs,
producing a `tbl_json` object. The `json_types()` function then inspects the
JSON associated with each row of the `tbl_json` object, and adds a new column
(`"type"`) that identifies the type according to the
[JSON standard](http://json.org/). We can then `count()` the number of
occurrences of each `name`, `type` combination to understand how many rows / 
observations contained these values.

::: {.callout-check-in}
10. Fill in the code below to discover how many different values are included in
the `companies` dataset?

```{r}
#| label: unique-values-in-dataset
#| eval: false

companies_values <- companies %>% 
  __________ %>% 
  json_types() %>% 
  __________
```

```{r}
#| label: values-in-data-soln
#| echo: false
#| eval: false

companies_values <- companies %>% 
  gather_object() %>% 
  json_types() %>% 
  count(name, type)
```

:::

Now that we know the name of **every** value in the `companies` dataset, we can
inspect how many of these columns were not captured when we used `spread-all()`.

::: {.callout-check-in}
11. Fill in the code below to uncover how many arrays from the `companies`
dataset were not spread into new columns.

```{r}
#| label: values-in-data-not-spread
#| eval: false

companies_values %>% 
  filter(
    ________ %in% ________
      ) %>% 
  filter(name != ____) 
```

```{r}
#| label: values-in-data-not-spread-soln
#| echo: false
#| eval: false

companies_values %>% 
  filter_out(
    name %in% colnames(companies_tidy)
    ) %>% 
  filter(name != "_id")
```
:::

### Step 3 -- Manually spread the problem values

Now that we know the names of the values that were not spread into columns, we
can wrangle specific values to capture their values. Let's do this with the
`acquisitions` value.

We do this with the following pipeline:

```{r}
#| eval: false
#| echo: true
#| label: process-for-manually-spreading-values

name_of_dataset %>% 
  enter_object(name_of_value) %>% 
  gather_array() %>% 
  spread_all()
```

The `enter_object()` function allows you to navigate a specific value of the
JSON data by referencing its name. `gather_array()` then collapses a JSON array
into index-value pairs, creating a new column `array.index` to store the index
of the array, and storing values in the 'JSON' attribute for further tidyjson
manipulation.

::: {.callout-check-in}
12. Fill in the code below to spread the `acquisitions` array of the `companies`
dataset. 

```{r}
#| label: spreading-acquisitions
#| eval: false

companies %>% 
  enter_object(____) %>% 
  gather_array() %>% 
  ________ 
```

13. What are the names of the new columns that appear in the spread dataset?
:::

### Step 4 -- Combining everything together

By combining with our initial top-level `spread_all()`, we can aggregate the
sale price of companies ('price_amount`) by `category_code` and the currency
it was sold in (`price_currency_code`). 

```{r}
#| label: mean-sale-price-for-each-category 
#| echo: true
#| message: false

company_max_sales <- companies %>% 
  spread_all() %>% 
  # Manually extracting information from acquisitions
  enter_object(acquisitions) %>% 
  gather_array() %>% 
  spread_all() %>% 
  # Removing missing values from relevant variables
  drop_na(category_code, price_currency_code, price_amount) %>% 
  # Finding max sale price for different company categories
  group_by(category_code, price_currency_code) %>%
  summarize(
    max_price = max(price_amount), 
    .groups = "drop"
            )

company_max_sales
```

To get a professional looking table, we can reformat the columns and use the 
**gt** package:

```{r}
#| label: nice-table-output
#| echo: true

library(gt)

company_max_sales |> 
  mutate(
    # Convert to dollar format
    max_price = scales::dollar(max_price), 
    # Remove _ from company sector names
    category_code = str_replace_all(category_code, 
                                    pattern = "_",
                                    replacement = " "), 
    # Capitalize the first letter of each word
    category_code = str_to_title(category_code)
    ) %>% 
  gt() %>% 
  cols_label(category_code = md("**Business Sector**"), 
             price_currency_code = md("**Currency of Sale**"), 
             max_price = md("**Max Company Sale Price**")
             )
```

<!-- ## STAT 541:  Step 5 -- Scaling to multiple columns -->

<!-- This is a powerful process, but you might be able to see how we don't -->
<!-- necessarily want to do this process repeatedly for every value `spread_all()` -->
<!-- was not able to parse. -->

<!-- ::: callout-tip -->
<!-- # Can you think of a way to use iteration to solve this task? -->
<!-- ::: -->

<!-- ```{r} -->
<!-- #| label: attempting-to-iterate -->
<!-- #| eval: false -->
<!-- #| echo: false -->

<!-- values_to_map <- companies_values %>%  -->
<!--   filter( -->
<!--     !name %in% colnames(companies_tidy) -->
<!--       ) %>%  -->
<!--   filter(name != "_id") %>%  -->
<!--   pull(name) %>%  -->
<!--   as.list() -->

<!-- manual_spread <- function(value){ -->
<!--   companies %>%  -->
<!--   enter_object({{value}}) %>%  -->
<!--   gather_array() %>%  -->
<!--   spread_all() -->
<!-- } -->

<!-- thing1 <- map(.x = values_to_map,  -->
<!--               .f = manual_spread -->
<!--               ) -->

<!-- reduce(thing1, .f = bind_cols) -->
<!-- ``` -->

# APIs 

Now that we've talked about the JSON format for data files, it's time we talked
about one of the primary vehicles for them: APIs.

::: {.callout-required-video}
This is the tip of the iceberg when it comes to APIs, but it will suffice to 
enable a lot of cool things for us. The video above used some nice examples like
the restaurant and travel services, but there are APIs for accessing all kinds
of data out there, like cat facts](https://alexwohlbruck.github.io/cat-facts/), 
[movie data](https://www.omdbapi.com/), and [NASA data](https://api.nasa.gov/).

</br>

::: youtube-video-container
{{< video https://www.youtube.com/watch?v=s7wmiS2mSXY >}}
:::
:::

For our class we're not interested in **creating** APIs, but mostly **using** 
them to access data that we otherwise couldn't get our hands on...at least not
as easily.

::: {.callout-required-reading}
[Dataquest: R API Tutorial](https://www.dataquest.io/blog/r-api-tutorial/)

**Warning!** The tutorial references the wrong link to access the API on the
number of people in space. The correct link is:
<http://api.open-notify.org/astros.json>
:::

::: {.callout-check-in}
11.  Which package does the tutorial use to make requests to the API server?

-   rvest
-   jsonlite
-   httr
-   tidyjson
-   XML

12.  What type of request will we likely start with when getting data from an
API?

-   `PULL()`
-   `GIMME()`
-   `GET()`
-   `SUMMON()`

13.  If your function call successfully requested information from the API, what
`status` should you see in the response object?

14. What `status` code would you get if your call to the API returned no content?
*Hint: The [http status codes](https://www.restapitutorial.com/httpstatuscodes)
might be helpful*
:::
