---
title: "Web Scraping"
editor: source
execute: 
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
```

After this lesson, you should be able to:

-   Use CSS Selectors and the Selector Gadget tool to locate data of interest
within a webpage
-   Use the `html_elements()` and `html_text()` functions within the `rvest`
packages to scrape data from webpage using CSS selectors

# Web scraping

We have talked about how to acquire data from APIs. Whenever an API is available
for your project, you should default to getting data from the API. Sometimes an
API will not be available, and **web scraping** is another means of getting
data.

Web scraping describes the use of code to extract information displayed on a web
page. In R, the `rvest` package (meant to sound like "harvest") offers tools for
scraping.

## Scraping ethics

### `robots.txt`

`robots.txt` is a file that some websites will publish to clarify what can and
cannot be scraped and other constraints about scraping. When a website publishes
this file, this we need to comply with the information in it for moral and legal
reasons.

We will look through the information in [this tutorial](https://www.zenrows.com/blog/robots-txt-web-scraping) and apply this
to the [cheese.com robots.txt file](https://www.cheese.com/robots.txt).

The [cheese.com](https://www.cheese.com/) `robots.txt` is very simple, it
says that anyone is allowed to scrape (`User-agent: *`). The file doesn't give
any additional instructions, such as `Crawl-delay` (how long you need to wait
between scraping each page), `Visit-time` (restrictions on time of day that
scraping is allowed), or `Request-rate` (restrictions on simultaneous requests).

### Further considerations

The article [Ethics in Web Scraping](https://carpentry.library.ucsb.edu/2022-05-12-ucsb-webscraping/06-Ethics-Legality-Webscraping/index.html) 
describes some good principles to ensure that we are valuing the labor that 
website owners invested to provide data and creating good from the information
we do scrape.

# HTML structure

HTML (hypertext markup language) is the formatting language used to create
webpages. Let's look at the core parts of HTML from the [rvest vignette](https://cran.r-project.org/web/packages/rvest/vignettes/rvest.html).

## Finding CSS Selectors

In order to gather information from a webpage, we must learn the language used
to identify patterns of specific information. For example, on the 
[alphabetical page of cheeses]](https://www.cheese.com/alphabetical/), we can
see that the data is represented in a consistent pattern of image + name of 
cheese.

We will identify data in a web page using a pattern matching language called
[CSS Selectors](https://css-tricks.com/how-css-selectors-work/) that can refer
to specific patterns in HTML, the language used to write web pages.

For example:

-   Selecting by tag:
    -   `"a"` selects all hyperlinks in a webpage ("a" represents "anchor" links
    in HTML)
    -   `"p"` selects all paragraph elements
    
-   Selecting by ID and class:
    -   `".description"` selects all elements with `class` equal to
    "description"
        -   The `.` at the beginning is what signifies `class` selection.
        -   This is one of the most common CSS selectors for scraping because in
        HTML, the `class` attribute is extremely commonly used to format webpage
        elements. (Any number of HTML elements can have the same `class`, which
        is not true for the `id` attribute.)
    -   `"#mainTitle"` selects the SINGLE element with **id** equal to
    "mainTitle"
        -   The `#` at the beginning is what signifies `id` selection.

``` html
<p class="title">Title of resource 1</p>
<p class="description">Description of resource 1</p>

<p class="title">Title of resource 2</p>
<p class="description">Description of resource 2</p>
```

::: {.callout-warning}
Websites change often! So if you are going to scrape a lot of data, it is
probably worthwhile to save and date a copy of the website. Otherwise, you may 
return after some time and your scraping code will include all of the wrong CSS
selectors.
:::

Although you can [learn how to use CSS Selectors by hand](%5BCSS%20Selectors%5D(https://css-tricks.com/how-css-selectors-work/)), 
we will use a shortcut by installing the 
[Selector Gadget](http://selectorgadget.com/) tool.

-   There is a version available for Chrome--add it to Chrome via the [Chome Web Store](https://chromewebstore.google.com/detail/selectorgadget/mhjhnkcfbdhnjickkkdbjoemdmbfginb).
    -   Make sure to pin the extension to the menu bar. (Click the 3 dots \> 
    Extensions \> Manage extensions. Click the "Details" button under
    SelectorGadget and toggle the "Pin to toolbar" option.)
-   There is also a version that can be saved as a bookmark in the browser--see [here](https://rvest.tidyverse.org/articles/selectorgadget.html).

::: {.callout-required-video}
::: youtube-video-container
<div style="padding:90.64% 0 0 0;position:relative;"><iframe src="https://player.vimeo.com/video/52055686?badge=0&amp;autopause=0&amp;player_id=0&amp;app_id=58479" frameborder="0" allow="autoplay; fullscreen; picture-in-picture; clipboard-write; encrypted-media; web-share" referrerpolicy="strict-origin-when-cross-origin" style="position:absolute;top:0;left:0;width:100%;height:100%;" title="SelectorGadget Screencast"></iframe></div><script src="https://player.vimeo.com/api/player.js"></script>
:::
:::


::: {.callout-check-in}
Head over to the [alphabetical page of the first 100 cheeses](https://www.cheese.com/alphabetical/?per_page=100).
Click the Selector Gadget extension icon or bookmark button. As you mouse over
the webpage, different parts will be highlighted in orange.

1.  What CSS selector would you use to obtain the image of the cheese?

<!-- img -->

2.  What CSS selector would you use to obtain the name of the cheese?

<!-- h3 -->

3.  What CSS selector would you use to obtain the link for each cheese?

<!-- .product-item a -->
:::

## Retrieving Data Using `rvest` and CSS Selectors {#sec-css-cheese}

Now that we have identified CSS selectors for the information we need, let's
fetch the data using the `rvest` package.

```{r}
#| label: read_html
#| echo: true
#| eval: false

library(rvest)

cheeses <- read_html("https://www.cheese.com/alphabetical/?per_page=100")
```

Once the webpage is loaded, we can retrieve data using the CSS selectors we
specified earlier. The code below retrieves three pieces of information about 
the cheeses on this website:

**1. name**

::: {.column-margin}
::: {.callout-tip}
# More Targetted CSS Selectors
I'm adding an `h3` to my CSS selector, since the store information is also
selected with the `.product-item`. If I don't grab the `h3` element, then 
the cheeses with store information will show up like this:
`"Stores >\n                \n            \n        \n    \n    Aged Gouda" `
:::
:::

```{r}
#| label: obtain-cheese-name
#| eval: false

cheeses %>%
  html_elements(".product-item") %>%
  html_element("h3") %>%
  html_text(trim = T) |> 
  head()
```

**2. HTML**

```{r}
#| label: obtain-cheese-url
#| eval: false

cheeses %>%
  html_elements(".product-item") %>%
  html_element("a") %>%
  html_attr("href") %>%
  url_absolute("https://www.cheese.com") |> 
  head()
```

::: {.column-margin}
::: {.callout-tip}
# Relative and Absolute URLs

The urls that are output by `html_attr("href")` are *relative* urls 
(e.g., `"/2-year-aged-cumin-gouda/"`) to get te *full* url, we need to paste the
base url (https://www.cheese.com) to the beginning of the relative url. The
`url_absolute()` function can do this for us. 

<!-- The function is essentially the same as: `str_c("https://www.cheese.com", .)`,  -->
<!-- where the string `"https://www.cheese.com"` is added to the beginning of every -->
<!-- relative url.  -->
:::
:::

**3. status of the image (exists or missing)**

```{r}
#| label: obtain-cheese-image
#| eval: false

cheeses %>%
  html_elements(".product-item") %>%
  html_element("a img") %>%
  html_attr("class") %>%
  head()
```

::: {.column-margin}
::: {.callout-tip}
# Detecting an Image

This seems like a great place to use `str_detect()`!
:::
:::

## Iteration

::: stat541
:::

Our goal is to obtain a dataset of characteristics about different cheeses 
from <https://www.cheese.com>. **Before doing any coding,** let's plan our
approach!

-   What functions will you write?
    -   We probably need a function to grab the text from the page.
    -   We can then use this function to grab different parts of an HTML page
    and piece them together
    
-   What arguments will they have?
    -   The function scraping the page should probably only take one input: the
    page
    -   The function grabbing text should have two inputs, (1) the page and (2)
    the CSS selector to use

-   How will you use your functions?
    -   These functions need to be iterated! We would need to call the function
    to scrape a page once for each page we want to scrape.

::: {.callout-check-in}
Let's carry out this plan to obtain the following information for **all**
cheeses in the database:

- cheese name
- URL for the cheese’s webpage (e.g., https://www.cheese.com/gouda/)
- whether or not the cheese has a picture (e.g., gouda has a picture, but bianco
does not).

To be kind to the website owners, we will add a 1 second pause between page
queries. (Note that you can view 100 cheeses at a time.)

4. All three of the CSS selectors from @sec-css-cheese, have something in
common. Can you spot it? Specifically, what input to `html_elements()` do they
all share?

<!-- .product-item -->

5. Now that we've found the similarities, let's spot the differences. Match each
CSS selector to where it is (exclusively) used. 

::: columns
::: {.column width="40%"}
- cheese name
- cheese URL
- cheese image

:::

::: {.column width="5%"}
:::

::: {.column width="55%"}
`h3`

`a`

`a img`

:::
:::


6. There are also differences in the HTML elements that are being extracted 
from the pages. Match each HTML extractor function to where it is used. 

::: columns
::: {.column width="40%"}
- cheese name
- cheese URL
- cheese image

:::

::: {.column width="5%"}
:::

::: {.column width="55%"}
`html_text()`

`html_attr()`

:::
:::

7. Let's use all this information to write a general function that can be used
for all three cases (name, URL, image). The function should take **four** 
arguments, (1) page to scrape, (2) css_selector, (3) node, and (4) what element
to extract. The following function calls should give you an idea of how the
function should operate.

```{r}
#| label: calling-scrape-function-name
#| eval: false

get_info_from_page(cheeses, 
                   css_selector = ".product-item", 
                   node = "h3 a") 

get_info_from_page(cheeses, 
                   css_selector = ".product-item", 
                   node = "a", 
                   extract = "href") |> 
  str_detect(pattern = "(#store-online-tabs)$")
  url_absolute("https://www.cheese.com")
```

**Complete the function below.**

```{r}
#| label: helper-function
#| eval: false

get_info_from_page <- function(page, css_selector, node, extract = "text") {
  
  # Grab the CSS and node elements from the page
  elements <- page %>%
    ________(css_selector) %>%
    ________(node)
  
  
  if (extract == "text") {
    # Extract the text from the element and trim it
    ________(elements, trim = TRUE)
  } else {
    # Extract what is specified
    html_attr(elements, extract)
  }
}
```

```{r}
#| label: helper-function-soln
#| echo: false

get_info_from_page <- function(page, css_selector, node, extract = "text") {
  
  # Grab the CSS and node elements from the page
  elements <- page %>%
    html_elements(css_selector) %>%
    html_element(node)
  
  
  if (extract == "text") {
    # Extract the text from the element and trim it
    html_text(elements, trim = TRUE)
  } else {
    # Extract what is specified
    html_attr(elements, extract)
  }
}
```

8. Our next step is to use this function in a larger `scrape_page()` function
that will obtain the necessary information for each page of cheeses. To be kind
to the website owners, let's add a 1 second pause between page queries.

**Fill in the code below to create the `scrape_cheese_page()` function.**

```{r}
#| label: scrape-page-function
#| eval: false

scrape_cheese_page <- function(url) {
    
    # 1 second crawl delay
    Sys.sleep(1)
    
    # Read the page
    page <- ________(url)
    
    # Grab cheese name
    cheese_names <- get_info_from_page(page, 
                   css_selector = ".product-item", 
                   node = ________) 
    
    # Grab cheese URL
    cheese_url <- get_info_from_page(page, 
                   css_selector = ".product-item", 
                   node = ________, 
                   extract = "href") |> 
      url_absolute("https://www.cheese.com")
    
    # Grab whether cheese has image
    cheese_image <- get_info_from_page(page, 
                   css_selector = ".product-item", 
                   node = ________, 
                   extract = "class") |> 
      # Detects if cheese has image
      str_detect("image-exists")
    

    #Put page elements into a data frame / tibble
    tibble(cheese = cheese_names,
           url = cheese_url,
           image = cheese_image
    )
}
```

```{r}
#| label: scrape-page-function-soln
#| echo: false

scrape_cheese_page <- function(url) {
    
    # 1 second crawl delay
    Sys.sleep(1)
    
    # Read the page
    page <- read_html(url)
    
    # Grab cheese name
    cheese_names <- get_info_from_page(page, 
                   css_selector = ".product-item", 
                   node = "h3 a") 
    
    # Grab cheese URL
    cheese_url <- get_info_from_page(page, 
                   css_selector = ".product-item", 
                   node = "a", 
                   extract = "href") |> 
      url_absolute("https://www.cheese.com")
    
    # Grab whether cheese has image
    cheese_image <- get_info_from_page(page, 
                   css_selector = ".product-item", 
                   node = "a img", 
                   extract = "class") |> 
      # Detects if cheese has image
      str_detect("image-exists")
    

    #Put page elements into a data frame / tibble
    tibble(cheese = cheese_names,
           url = cheese_url,
           image = cheese_image
    )
}
```

9. Finally, let's iterate over **all** the pages of cheese! We can figure out
how many pages we need to scrape over by grabbing the number of pages from the 
bottom of the [first alphabetical page (per page = 100)](https://www.cheese.com/alphabetical/?per_page=100).
This method is a bit less error prone than manually typing `21`!

```{r}
#| eval: false

# A less error prone 
total_pages <- cheeses %>%
  html_elements(".page-link:nth-child(10)") %>%
  html_text()
```

**Fill in the code below to `map()` over _every_ page of cheese.**

```{r}
#| label: map-scrape
#| eval: false

base_url <- "https://www.cheese.com/alphabetical/?per_page=100&page="

urls_all_pages <- str_c(base_url, 1:______)

all_pages <- map(.x = ________, 
                 .f = ________)
```

```{r}
#| label:  map-scrape-soln
#| echo: false
#| cache: true
#| eval: false

base_url <- "https://www.cheese.com/alphabetical/?per_page=100&page="

urls_all_pages <- str_c(base_url, 1:total_pages)

all_pages <- map(.x = urls_all_pages, 
                 .f = scrape_cheese_page)
```

10. The `map()` function returns a list of dataframes. What function should 
we use to bind each of these dataframes together to form **one** large dataframe
on cheeses?

- `bind_cols()`
- `bind_rows()`
- `inner_join()`
- `full_join()`

```{r}
#| label: bind-lists
#| echo: false
#| eval: false

all_cheeses <- bind_rows(all_pages)
all_cheeses
```
:::
