---
title: "Week 6 - Making Maps!"
format: 
  revealjs:
    theme: [style.scss, simple]
    auto-stretch: false
editor: source
---

```{r}
library(tidyverse)
```


# Let's explore some different ways to make maps!

# Option 1: Latitudes & Longitudes

## California Counties

```{r}
#| echo: true
#| code-line-numbers: false

ca_counties <- map_data("county", "california")
```

::: {.midi}
```{r}
ca_counties
```
:::

. . .

Notice Alameda has **multiple** rows in these data! What do you think each
unique `long` and `lat` correspond to?

## How would you expect this plot to look?

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: false

ggplot(data = ca_counties, 
       mapping = aes(long, lat)
       ) + 
  geom_point(size = 0.25)
```

. . .

```{r}
#| out-width: 80%
#| fig-align: center

ggplot(data = ca_counties, 
       mapping = aes(long, lat)
       ) + 
  geom_point(size = 0.25) 
```

## That plot seems a bit off!

::: {.midi}
```{r}
#| echo: true
#| code-line-numbers: false
#| out-width: 80%
#| fig-align: center

ggplot(data = ca_counties, 
       mapping = aes(long, lat)
       ) + 
  geom_point(size = 0.25) +
  coord_sf()
```
:::


## Now suppose I used `geom_polygon()` instead

How would you expect this plot to look?

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: false

ggplot(data = ca_counties, 
       mapping = aes(long, lat, group = group)
       ) +
  geom_polygon(fill = "white", colour = "grey50") + 
  coord_sf()

```

## Getting Shapes

```{r}
#| echo: false
#| eval: true

ggplot(data = ca_counties, 
       mapping = aes(long, lat, group = group)
       ) +
  geom_polygon(fill = "white", colour = "grey50") + 
  coord_sf()
```

# Option 2: Shape Files

## The `tigris` Package

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: false

library(tigris)

ca_counties <- tigris::counties(state = "california")
```

</br>

::: {.midi}
```{r}
#| eval: true
#| echo: false

# Because tigris leaves a message everytime the code is run!
ca_counties <- tigris::counties(state = "california", progress_bar = FALSE)

ca_counties |> 
  glimpse()
```
:::

## Using Shape Files

::: {.midi}
```{r}
#| label: shapes-with-coord-sf
#| echo: true
#| code-line-numbers: false
#| out-width: 80%

ggplot(data = ca_counties) +
  geom_sf(fill = "white", 
          color = "grey25", 
          linewidth = 0.1) +
  theme_void()
```
:::

## Why not use `leaflet`?

```{r}
#| label: convert-data-for-leaflet
#| echo: false

library(sf)
ca_counties <- st_transform(ca_counties, 4326)
```

::: {.midi}
```{r}
#| echo: true
#| code-line-numbers: false
#| output-location: column

library(leaflet)

leaflet(ca_counties) %>%
  addPolygons(
    fillColor = "white",
    fillOpacity = 1,
    color = "grey25",
    weight = 0.1
  )
```
:::
