---
title: "Adverbs with `purrr`"
format: 
  revealjs:
    theme: [style.scss, simple]
    auto-stretch: false
editor: source
---

```{r}
library(tidyverse)
```


## What is an adverb?

* In English, an adverb *modifies* a verb, e.g. "He ate" vs "He ate **loudly**".  

* In tidyverse, an adverb *modifies* a function.

* Syntax example: Notice that parentheses are around `mean` not around `mean(1:10)`

```{r}
#| eval: false
#| echo: true
safely(mean)(1:10)
```

## `safely()`

* The `safely()` adverb allows a function to run and return an error without breaking the code.

* It causes the function to return two objects:
    + `result`: containing whatever the function would normally return; or, `NULL` if there was an error.
    + `error`: TRUE or FALSE and the error message
    
    
## `safely()`

```{r}
#| error: true
#| echo: true
mean(1:10)
safely(mean)(1:10)
```

## `safely()`

```{r}
#| error: true
#| echo: true
mean("a" + 1)
safely(mean)("a" + 1)
```

## `safely()`

* `safely()` is particularly useful for webscraping and API calls

* Use it when you aren't sure if your `GET` or `fromJSON` or similar function is going to succeed.

* Then, check the `$error` object to determine if you should return an `NA` or actual results.

* Remember to access the `$result` object when you need the actual results!


## Other handy adverbs:

* `possibly()`: Same as to `safely()`, but you can have it return `NA` or similar as a result if the function errors.

* `quietly()`: Keeps messages from printing, but still stores them.

* `insistently()`: Retries a function multiple times if it fails, allowing you to wrap unstable operations (e.g., API requests) to work reliably.

* `slowly()`: Forces a function to pause between executions, useful for rate-limiting calls.