---
title: "Writing Efficient Code"
editor: source
execute: 
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(tictoc)
library(microbenchmark)

options(digits = 2)
```

In this module, you'll learn a bit about how to make your code faster in general.


# Code speed

Fortunately, we stand on the shoulders of giants. Many skilled computer
scientists have put a lot of time and effort into writing R functions that run
as fast as possible. For you, most of the work to speed up code is simply 
finding the right packages to rely on.

However, there are a few baseline principles that can get you a long way. If
your code takes a long time to run, the reason is often one of these:

-   You are doing a very large computation, relative to the power of your
computer.
-   You are repeating a slow process many times.
-   You are trying to do simple things, but on a very large object.

To speed up the code, without deep knowledge of computer algorithms and inner
workings, you can sometimes come up with clever ways to avoid these pitfalls.

First, though: as you start thinking about writing faster code, you'll need a
way to check whether your improvements actually sped up the code.

## 📖 Required Reading: [The tictoc and microbenchmark packages](https://www.jumpingrivers.com/blog/timing-in-r/)

## 📖 Required Reading: [The bench package](https://www.tidyverse.org/blog/2018/06/bench-1.0.1/)

## 📖 Required Reading: [`lobstr::obj_size()`](https://lobstr.r-lib.org/reference/obj_size.html)

# Tip 1: Avoid larger calculations and memory demands

## Save smaller data frames, if you are going to use them many times

Consider the following dataset:

```{r}
#| echo: true
#| message: false
#| label: load-data

library(fivethirtyeight)

glimpse(congress_age)
```

Suppose we want to do some exploratory analysis on these data. Let's use the
`tic()` and `toc()` functions from the **tictoc** packages to check how much
time it takes to run this code:

```{r}
#| results: false
#| message: false
#| fig-show: hide
#| echo: true
#| label: exploratory-analysis

tic()

congress_age %>%
  filter(chamber == "senate") %>%
  summarize(mean(age))

congress_age %>%
  filter(chamber == "senate") %>%
  ggplot() + 
  geom_histogram(mapping = aes(x = age))

congress_age %>%
  filter(chamber == "senate") %>%
  slice_max(n = 5, order_by = age)

```

```{r}
#| echo: true
toc()
```

These are all three reasonable things to do, and they can't be done in the same pipeline. But wait - we just made R do the process of subsetting to only Senators three separate times!

```{r, results = FALSE}
#| echo: true

tic()
  congress_age %>%
    filter(chamber == "senate")
```

```{r}
#| echo: true
toc()
```

Instead, how about we filter the data once, save it into an object, and call on that object:

```{r}
#| echo: true
#| fig-show: hide
#| message: false
#| results: false

tic()

 senate <- congress_age %>%
  filter(chamber == "senate")

senate %>%
  summarize(mean(age))

senate %>%
  ggplot() + 
  geom_histogram(mapping = aes(x = age))


senate %>%
  slice_max(n = 5, order_by = age)
```

```{r}
#| echo: true

toc()
```

That is **half** the run time! Yes, we are dealing with minute increments of
time (less than a second), but hopefully you could see how this might scale to
larger calcuations. 

# Tip 2: Avoid repeating slow steps

## Avoid for-loops

For-loops are deathly slow in R. If you absolutely must iterate over a process,
rely on the `map()` or `apply()` function families. 

::: {.callout-note}
# Why choose `map()` instead of `apply()`? 

The `map()` and `apply()` functions will typically be close to the same speed, 
so the largest  argument in favor of the `map()` family of functions are (1)
their consistent syntax, and (2) their consistent object output. 

The `map()` function **always** returns a list and the `map_XXX()` variants
**always** return a vector of a specified data type. Whereas, `apply()` returns
a vector or a list, `lapply()` returns a list, and `sapply()` returns a vetor or
a matrix. 
:::

Here is a motivating example demonstrating the difference in run time between 
a function which loops over every column in a data frame to extract the data
type and a function which uses `map_chr()` to apply the `class()` function to
every column of a data frame:

```{r}
#| echo: true
#| cache: true
#| warning: false
#| label: compare-loops-map-apply

loop_func <- function(df){
  
  # initialize the type object
  type <- rep(NA, ncol(df))
  
  for(i in 1:ncol(df)){
    
    # store the class of the ith column of df into the ith entry of type 
    type[i] <- class(df[,i])
  }
  
  return(type)
}

map_func <- function(df){
  
  type <- map_chr(df, class)
  
  return(type)

  }

dat <- as.data.frame(
  matrix(1,
         nrow = 5,
         ncol = 100000)
  )

microbenchmark::microbenchmark(loop_func(dat),
                               map_func(dat),
                               times = 3)
```

That's a pretty big difference!

## Use vectorized functions

Better even than `map()` or `apply()` is not to iterate at all! Writing 
vectorized functions is tough, but do-able in many cases. Here's an example:

## 🎥 Required Video: [Simple Vectorizing Example](https://www.youtube.com/embed/ShewFCbPHv8)

Better yet - rely on built-in functions in other packages to do it for you!

## Allocate memory in advance

A trick we use very often in coding is to "tack on" values to the end of a
vector or list as we iterate through something. However, this is actually very
slow! If you know how long your list is going to end up, consider pre-creating
that list.

```{r}
#| echo: true

do_stuff_allocate <- function(reps) {
  results <- vector(length = reps)
  for (i in 1:reps) {
    results[i] <- i
  }
  return(results)
}

do_stuff_tackon <- function(reps) {
   results <- c()
  for (i in 1:reps) {
    results <- c(results, i)
  }
  return(results)
}

microbenchmark::microbenchmark(
  do_stuff_allocate(1000),
  do_stuff_tackon(1000)
)

```

# Tip 3: Use faster existing functions

Because R has so many packages, there are often many functions to perform the
same task. Not all functions are created equal!

## `data.table`

For speeding up work with data frames, no package is better than `data.table`!

## 📖 Required Video: [5-minute intro to data.table](https://www.youtube.com/embed/uueVddWwbkk)

## 📖 Required Reading: [data.table cheatsheet](https://pbs.twimg.com/media/Dwoq0ZQXQAE5lJt.jpg)

::: {.callout-important}
# Check-in: `data.table`

**Question 1:** Here is some `dplyr` code. Re-write it in `data.table` syntax. 
Then perform a speed test to see which one is faster, and by how much.

```{r}
#| echo: true
#| eval: false

congress_age %>%
  group_by(state) %>%
  summarize(mean_age = mean(age))

```
:::

# Tip 4: Only improve what needs improving

> “We should forget about small efficiencies, say about 97% of the time:
> premature optimization is the root of all evil. Yet we should not pass up our
> opportunities in that critical 3%. A good programmer will not be lulled into
> complacency by such reasoning, \[they\] will be wise to look carefully at the
> critical code; but only after that code has been identified.” — Donald Knuth.

Speed and computational efficiency are important, but there can be a tradeoff. 
If you make a small speed improvement, but your code becomes overly complex, 
confusing, or difficult to edit and test it's not worth it!

Also, consider your time and energy as the programmer. Suppose you have a 
function that takes 30-minutes to run. It relies on a subfunction that takes 
30-seconds to run. Should you really spend your efforts making the subfunction
take only 10-seconds?

The art of finding the slow bits of your code where you can make the most
improvement is called **profiling**.

## 📖 Required Reading: [Profiling] [](http://adv-r.had.co.nz/Profiling.html#profiling)

## 📖 Recommended Reading: [Measuring Performance](https://adv-r.hadley.nz/perf-measure.html)

::: {.callout-tip}
# Skimming is sufficient!

For this reading, I strongly encourage you to skim. The code aspects are mostly
more complicated that we expect in this class; but the general overview of the
principles behind speeding up code is helpful.

In your work, you will most likely "profile" manually, by `tictoc`-ing bits of
your code and so forth, not by these fancy methods.
:::

::: {.callout-important}
# Check-in: Profiling

**Question 2:** Why is `vapply()` faster than `sapply()`?

a.  It specifies the output type.
b.  It specifies the input type.
c.  It performs fewer safety checks.
d.  It performs fewer iterations.

**Question 3:** Here are two other ways to compute the square root of a vector.
Which is faster?  fastest? Use microbenchmarking to test your answers!

```{r}
#| echo: true
#| eval: false
x ^ (1 / 2)
exp(log(x) / 2)
```

**Question 4:** Why is `mean(x)` slower than `sum(x)/length(x)`?

a.  `sum()` is vectorized, while `mean()` is not.
b.  A mean is a smaller number than a sum, so it needs less memory.
c.  `mean()` calculates the mean twice just in case.
d.  `mean()` was written in an old version of R.

:::

## Optional: Super advanced mode

The following are some suggestions of ways you can level-up your code speed. These are outside the scope of this class, but you are welcome to explore them!

-   [Dive deep into the R efficiency rabbit hole...](https://csgillespie.github.io/efficientR/)
-   [Parallelize your computations](http://zevross.com/blog/2019/02/12/dramatically-speed-up-your-r-purrr-functions-with-the-furrr-package/)
-   [Make your code run in C++ with Rcpp](http://adv-r.had.co.nz/Rcpp.html#rcpp)
-   Learn to use [recursion](https://techvidvan.com/tutorials/recursion-in-r/)
-   Learn more about [memory allocation](https://adv-r.hadley.nz/names-values.html) and [garbage collecting](http://adv-r.had.co.nz/memory.html) in R.
