---
title: "Iterating Over Functions"
editor: source
---

In this unit, you’ll learn tools for iteration---repeatedly performing the same
action on different objects. Iteration in R generally tends to look a bit
different from other programming languages. Much of iteration we get for free! 
For example, if you want to double a numeric vector `x` in R, you can just 
write `2 * x`, whereas in many other languages you would need to explicitly 
double each element of `x` using some sort of `for` loop.

In R, there are generally two methods for iteration---`for()` loops and 
functionals. We will start with a review of `for()` loops before hopping over
to functionals. 

```{r}
#| label: packages
#| include: false
library(tidyverse)
```

# `for` loops

In R, `for` loops have the following general structure:

```{r for_syntax, eval=FALSE}
for (i in some_vector) {
    # Code to do stuff with i
}
```

`some_vector` can be **any** vector, including:

- An indexing vector: `1:3`
- A character vector: `c("group1", "group2", "group3")`
- A vector of any other class

```{r}
#| label: for-examples

groups <- c("group1", "group2", "group3")

for (i in 1:3) {
    print(groups[i])
}

for (g in groups) {
    print(g)
}
```

The `seq_along()` function generates an integer sequence from 1 to the length of
the vector supplied. A nice feature of `seq_along()` is that it generates an
empty iteration vector if the vector you're iterating over itself has length 0.

```{r}
#| label: for-seq-along

seq_along(groups)
no_groups <- c()
seq_along(no_groups)

for (i in seq_along(groups)) {
    print(groups[i])
}

for (i in seq_along(no_groups)) {
    print(no_groups[i])
}
```

Closely related to `seq_along()` is `seq_len()`. While `seq_along(x)` generates
an integer sequence from 1 to `length(x)`, `seq_len(x)` takes `x` itself to be a
length:

```{r}
#| label: for-seq-len

seq_len(3)
seq_len(0)

for (i in seq_len(length(groups))) {
    print(groups[i])
}

for (i in seq_len(length(no_groups))) {
    print(no_groups[i])
}
```

`seq_len()` is useful for iterating over the rows of a data frame because `seq_along()` would iterate over columns:

```{r}
#| label: seq-len-df

small_data <- tibble(a = 1:2, 
                     b = 2:3, 
                     c = 4:5)

for (col in small_data) {
    print(col)
}

for (r in seq_len(nrow(small_data))) {
    print(r)
}
```

Often we'll want to store output created during a `for` loop. We can create
storage containers with the `vector()` function:

```{r}
#| label: store-results

char_storage <- vector("character", 
                       length = 3)
char_storage

num_storage <- vector("numeric", 
                      length = 3)
num_storage

list_storage <- vector("list", 
                       length = 3)
list_storage

for (i in seq_len(3)) {
    char_storage[i] <- str_c("Number: ", i)
    num_storage[i] <- 2*i
    list_storage[[i]] <- i # Note the [[ for subsetting here
}

char_storage
num_storage
list_storage
```

### Exercises

Write `for()`-loops that do each of the following:

1. Prints the even numbers from 1:20. 
  - Produce the same output with the `seq()` function!

<details>
    <summary>Solutions</summary>

```{r}
#| label: for-ex-solutions-1

for (i in seq_len(10)) {
    print(2*i)
}

seq(from = 2, to = 20, by = 2)
```

</details>

2. Iterates over the `month.name` vector (built-in to base R) and stores a
character vector of output containing strings like "Month 1: January", "Month 2:
February".
  - Produce the same output with `str_c()` only! 

<details>
    <summary>Solutions</summary>

```{r}
#| label: for-ex-solutions-2

month_strings <- vector("character", length = length(month.name))

for (i in seq_along(month.name)) {
    month_strings[i] <- str_c("Month ", i, ": ", month.name[i])
}
month_strings

str_c("Month ", 1:12, ": ", month.name)
```

</details>

3. Store the `class()` (type) of every column in the `mtcars` data frame. 

<details>
    <summary>Solution</summary>

```{r}
#| label: col_classes_soln

col_classes <- vector("character", 
                      ncol(mtcars)
                      )

# Data frames are **lists** of columns, so this loop iterates over the columns
for (i in seq_along(mtcars)) {
    col_classes[i] <- class(mtcars[[i]])
}
```

</details>

# Iteration with Functionals

> A functional is a function that takes a function as an input and returns a
> vector as output. 
> - Hadley Wickham

## 📖 Required Reading: [Functionals](https://adv-r.hadley.nz/functionals.html)

`purrr` is a `tidyverse` package that provides several useful functions for
iteration. The main advantages of `purrr` include:

- Improved readability of R code
- Reduction in the "overhead" in writing a `for` loop (creating storage
containers and writing the `for (i in ...)`)

In `purrr`, we can use the family of `map()` functions to apply a function to
each element of a list or vector. We can think of this as mapping an input (a
list or vector) to a new output via a function. Let's look at the [`purrr` cheatsheet](https://rstudio.github.io/cheatsheets/purrr.pdf) to look at
graphical representations of how these functions work.

- `map()` returns a list
- `map_chr()` returns a character vector
- `map_lgl()` returns a logical vector
- `map_int()` returns an integer vector
- `map_dbl()` returns a numeric vector
- `map_vec()` returns a vector of a different (non-atomic) type (like dates)

To get the `class()` of each data frame column, `map_chr()` is sensible because
classes are strings:

```{r}
#| label: purrr_col_classes

map_chr(mtcars, .f = class)
```

Let's get the class of each variable in `diamonds`:

```{r}
#| label:  purrr_col_classes_error
#| error: true

map_chr(diamonds, class)
```

What happened!? `map_chr()` was expecting to create a character vector with one
element per element (column) in `diamonds`. But something happened in column 2
with the `cut` variable. Let's figure out what happened:

```{r}
#| label: class_diamonds_cut

class(diamonds$cut)
```

Ah! `cut` has **two** classes. In this case, `map()` (which returns a list) is
the best option because some variables have multiple classes:

```{r}
#| label: purrr_col_classes_no_error

map(diamonds, class)
```

The error we encountered with `map_chr()` is a nice feature of `purrr` because
it allows us to be very sure of the type of output we are getting. Failing
loudly is vastly preferable to getting unexpected outputs silently because we
can catch errors earlier!

We can combine `map_*()` functions with tidy selection for some powerful
variable summaries that require much less code than `for()` loops.

```{r}
#| label:  purrr_examples_by_col

diamonds %>% 
  select(where(is.numeric)) %>% 
  map_dbl(.f = mean)

diamonds %>% 
  select(!where(is.numeric)) %>% 
  map_int(.f = n_distinct)
```

### Exercises

4. Write a function called `summ_stats()` that takes a numeric vector `x` as
input and returns the mean, median, standard deviation, and IQR as a data frame.
You can use `tibble()` to create the data frame.
    * Example: `tibble(a = 1:2, b = 2:3)` creates a data frame with variables
    `a` and `b`.

<details>
    <summary>Solution</summary>

```{r}
#| label: ex_summ_stats_soln1

summ_stats <- function(x) {
    tibble(
        mean = mean(x, na.rm = TRUE),
        median = median(x, na.rm = TRUE),
        sd = sd(x, na.rm = TRUE),
        iqr = IQR(x, na.rm = TRUE)
    ) 
}
```

</details>

5. Use `map()` to apply your `summ_stats()` function to the numeric columns in 
the `diamonds` dataset. 

::: columns
::: {.column width="50%"}
::: {.callout-tip}
Look up the `bind_rows()` documentation from `dplyr` to combine summary
statistics for all quantitative variables into one data frame. The `.id` 
argument will be especially helpful in adding the variable names! 
:::
:::

::: {.column width="3%"}
:::

::: {.column width="47%"}
::: {.callout-important}
# Consider pivoting !

The output would be easier for the reader to understand if the summary
statistics were rows and the variable names were columns.
:::
:::
:::

<details>
    <summary>Solution</summary>

```{r}
#| label: ex_summ_stats_soln2

diamonds_num <- diamonds %>% 
  select(where(is.numeric))

map(diamonds_num, .f = summ_stats) %>% 
  bind_rows(.id = "variable") %>% 
  pivot_longer(cols = -variable, 
               names_to = "statistic", 
               values_to = "value") %>% 
  pivot_wider(names_from = variable, values_from = value)

```

</details>

6. Write a `for()` loop to achieve the same result. Which do you prefer in
terms of ease of code writing and readability?

<details>
    <summary>Solution</summary>

```{r}
#| label: ex_summ_stats_soln3

diamonds_summ_stats2 <- vector("list", length = ncol(diamonds_num))

for (i in seq_along(diamonds_num)) {
    diamonds_summ_stats2[[i]] <- summ_stats(diamonds_num[[i]])
}

diamonds_summ_stats2 %>%
    set_names(colnames(diamonds_num)) %>%
    bind_rows(.id = "variable")
```

</details>

## Multiple Inputs

`purrr` also offers the `pmap()` family of functions that take multiple inputs
and loops over them simultaneously. Let's look at the [`purrr` cheatsheet](https://rstudio.github.io/cheatsheets/purrr.pdf) to look at
graphical representations of how these functions work.

```{r}
#| label: pmap_ex1

string_data <- tibble(
    string = c("apple", "banana", "cherry"),
    pattern = c("p", "n", "h"),
    replacement = c("P", "N", "H")
)
string_data

pmap_chr(string_data, .f = str_replace_all)
```

Note how the column names in `string_data` exactly match the argument names in
`str_replace_all()`. The iteration that is happening is across rows, and the
multiple arguments in `str_replace_all()` are being matched by name.

We can also use `pmap()` to specify variations in some arguments but leave some
arguments constant across the iterations:

```{r}
#| label: pmap_ex2

string_data <- tibble(
    pattern = c("p", "n", "h"),
    replacement = c("P", "N", "H")
)

pmap_chr(string_data, str_replace_all, string = "ppp nnn hhh")
```

### Exercises

7. Create 2 small examples that show how `pmap()` works with `str_sub()`. Your
examples should:

- Use different arguments for `string`, `start`, and `end`
- Use different arguments for `start` and `end` but a fixed `string`

<details>
    <summary>Solution</summary>

```{r}
#| label: pmap_ex_str_remove_soln

string_data <- tibble(
    string = c("apple", "banana", "cherry"),
    start = c(1, 2, 4),
    end = c(2, 3, 5)
)

pmap_chr(string_data, str_sub)

string_data <- tibble(
    start = c(1, 2, 4),
    end = c(2, 3, 5)
)
pmap_chr(string_data, str_sub, string = "abcde")
```

</details>