---
title: "Hierarchical Data"
format: html
editor: source
execute: 
  echo: false
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(jsonlite)
```

# Tabular Data

It's likely that a huge portion of the datasets you've worked with in your life
thus far are very well structured and organized to the point of being
**tabular**. That is, the dataset consists of a single table with rows and
columns. If you're not familiar with the concept of **tidy data** I suggest you
check out this [`tidyr` vignette](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html)
on what it means for a dataset to be "tidy." Most of your datasets have likely
even been tidy:

-   Each variable forms a column
-   Each observation forms a row

Almost everyone's experiences beyond this probably involved taking multiple
**tidy** datasets and combining them in some way, but the overall dataset still
fit into this tabular structure. These often came in the form of TXT, CSV,
Excel, or TSV files.

Our world is rich with data and a great deal of it doesn't come in nice, neat
tables...BUT a lot of it is still extremely structured and well organized.

## Structured, but Non-Tabular Data

You've likely already noticed, via your work with `leaflet`, there is another
very popular data format called "JSON". JSON is short for **J**ava**S**cript
**O**bject **N**otation and is a syntax for storing and exchanging data...but
data in a much broader sense than we're used to thinking about.

In the middle of a statistics-related course when I hear the word "data" my mind
instinctively envisions something tabular with which we could do some sort of
visualization or analysis with. However, in the broadest sense, the word "data"
really just describes pieces of information...which encapsulates everything from
movie theater schedules when you look up showtimes on a website to what your
cell phone *uses* as part of your "data plan."

While the website for your favorite movie theater may not have had your
visualization and analysis plans in mind, their data may still be of interest
to us. Many companies and organizations have built tools that allow access to
their data in a more streamlined way, and JSON has become a ubiquitous format
for many of these data sources to use.

# Hierarchical Data

One of the reasons the JSON format is so important and useful is because it can
accommodate very complex types and shapes of data. So while it may not always be
workable into a tabular form, it still maintains a high degree of structure that
we can exploit!

## 📖 Required Reading: [*R for Data Science*, Chapter 23](https://r4ds.hadley.nz/rectangling.html)

::: {.callout-note}
# Only read Sections 1-4 for now!
:::

## Check-in: Unnesting

1. What happens when you use `unnest_wider()` with unnamed list-columns like 
`df2`? 

(a) What argument is now necessary? 

<!-- ! Can't unnest elements with missing names. -->
<!-- ℹ Supply `names_sep` to generate automatic names. -->

(b) How does `unnest_wider()` reconcile columns with different lengths?

```{r}
#| include: false
#| label: code-for-q1

df2 <- tribble(
  ~x, ~y,
  1, list(11, 12, 13),
  2, list(21),
  3, list(31, 32),
)

unnest_wider(df2, col = y, names_sep = "_")

```

2. From time-to-time you encounter data frames with multiple list-columns with
aligned values. For example, in the following data frame, the values of `y` and
`z` are aligned (i.e. `y` and `z` will always have the same length within a row,
and the first value of `y` corresponds to the first value of `z`). 

```{r}
#| include: false
#| label: code-for-q2

df4 <- tribble(
  ~x, ~y, ~z,
  "a", list("y-a-1", "y-a-2"), list("z-a-1", "z-a-2"),
  "b", list("y-b-1", "y-b-2", "y-b-3"), list("z-b-1", "z-b-2", "z-b-3")
)

```

(a) What happens if you apply two `unnest_longer()` calls to this data frame? 

```{r}
#| include: false
#| label: code-for-q2-a

unnest_longer(df4, col = y) %>% 
  unnest_longer(col = z)
```

(b) How can you preserve the relationship between x and y? *Hint: Carefully
read the documentation*

```{r}
#| include: false
#| label: code-for-q2-b

unnest_longer(df4, col = c(y, z))
```

# JSON Data

## 📖 Required Reading: [*R for Data Science*, Chapter 23](https://r4ds.hadley.nz/rectangling.html)

::: {.callout-note}
# Now read Sections 5 & 6! 
:::

## Check-in: JSON, pt. 1

1. Objects are held in \_\_\_\_\_.

-   parentheses
-   double quotes
-   square brackets
-   curly braces

2.  While it may seem tedious compared to our usual tabular forms of data, JSON
data is easy to read because it \_\_\_\_\_.

-   comes in name / value pairs
-   is separated by "-----"
-   comes in table / name/value triplets

3.  Create rectangular versions of `df_col` and `df_row` (shown below). 

```{r}
#| eval: false
#| echo: true
#| label: code-for-q3

json_col <- parse_json('
  {
    "x": ["a", "x", "z"],
    "y": [10, null, 3]
  }
')

json_row <- parse_json('
  [
    {"x": "a", "y": 10},
    {"x": "x", "y": null},
    {"x": "z", "y": 3}
  ]
')

df_col <- tibble(json = list(json_col)) 

df_row <- tibble(json = json_row)

```

```{r}
#| eval: false
#| echo: false
#| label: solution-for-q3

df_col %>% 
  unnest_wider(col = json) %>% 
  unnest_longer(col = c(x, y))

df_row %>% 
  unnest_wider(col = json)
```

# Working with JSON Data

The chapter covers a few functions from the `jsonlite` package, but there are 
several other packages for working with JSON data in R (e.g., `rjson`, 
`rjsonio`). These packages automatically transform JSON data into nested R
lists or complex data frames. However, working with these complex objects can be
difficult.

The `tidyjson` package takes a different approach to structuring JSON data into
tidy data frames. Similar to `tidyr`, `tidyjson` builds a grammar for
manipulating JSON into a tidy table structure.

`tidyjson` is based on the following principles:

-   Work on a single JSON document, or on a collection of related documents
-   Create pipelines with the pipe operator (`%>%`), producing code that can be
read from left to right
-   Guarantee the structure of the data produced, even if the input JSON
structure changes (with the exception of `spread_all()`)
-   Work with arbitrarily nested arrays or objects
-   Handle 'ragged' arrays and / or objects (varying lengths by document)
-   Ensure edge cases are handled correctly (especially empty data)
-   Integrate seamlessly with `dplyr`, allowing `tbl_json` objects to pipe in
and out of `dplyr` verbs where reasonable

## Check-in: JSON, pt. 2

Let's use the `companies` JSON dataset from the **tidyjson** package to explore
tools from the **tidyjson** package. This exploration parallels the `worldbank`
examples from the [**tidyjson** vignette](https://cran.r-project.org/web/packages/tidyjson/vignettes/introduction-to-tidyjson.html).

**Step 1** -- Use `spread_all()` to spread the values of a JSON object

```{r}
#| label: explore-worldbank-data
#| message: false

library(tidyjson)
companies_tidy <- spread_all(companies)
```

5.  When you use the `spread_all()` function on the `companies` dataset, how
many columns are created?

<!-- `r ncol(companies_tidy)` -->

**Step 2** -- Find the values that were not able to be spread

To identify the names of an object, their type, and frequency we use the
following pipeline:

```{r}
#| eval: false
#| echo: true
#| label: process-for-finding-missed-values

name_of_data %>% 
  gather_object() %>% 
  json_types() %>% 
  count(name, type)
```

The `gather_object()` function collapses a JSON object into name-value pairs,
producing a `tbl_json` object. The `json_types()` function then inspects the
JSON associated with each row of the `tbl_json` object, and adds a new column
(`"type"`) that identifies the type according to the
[JSON standard](http://json.org/). We can then `count()` the number of
occurrences of each `name`, `type` combination to understand how many rows / 
observations contained these values.

6.  How many different values are included in the `companies` dataset?

```{r}
#| label: unique-values-in-dataset
#| eval: false

companies_values <- companies %>% 
  gather_object() %>% 
  json_types() %>% 
  count(name, type)
```

Now that we know the name of **every** value in the `companies` dataset, we can
inspect how many of these columns were not captured when we used `spread-all()`.

7.  How many arrays from the `companies` dataset were not spread into new
columns?

```{r}
#| label: values-in-data-not-spread
#| eval: false

companies_values %>% 
  filter(
    ! name %in% colnames(companies_tidy)
      ) %>% 
  filter(name != "_id") 
```

**Step 3** -- Manually spread the problem values

Now that we know the names of the values that were not spread into columns, we
can wrangle specific values to capture their values. Let's do this with the
`acquisitions` value.

We do this with the following pipeline:

```{r}
#| eval: false
#| echo: true
#| label: process-for-manually-spreading-values

name_of_dataset %>% 
  enter_object(name_of_value) %>% 
  gather_array() %>% 
  spread_all()
```

The `enter_object()` function allows you to navigate a specific value of the
JSON data by referencing its name. `gather_array()` then collapses a JSON array
into index-value pairs, creating a new column `array.index` to store the index
of the array, and storing values in the 'JSON' attribute for further tidyjson
manipulation.

8.  Using this pipeline on the `acquisitions` array of the `companies` dataset,
what values are stored in this array? i.e., what are the names of the new
columns that appear in the spread dataset?

```{r}
#| label: spreading-acquisitions
#| eval: false

companies %>% 
  enter_object(acquisitions) %>% 
  gather_array() %>% 
  spread_all() 
```

**Step 4** -- Combining everything together

By combining with our initial top-level `spread_all()`, we can aggregate the
sale price of companies by `category_code` and the currency it was sold in.

```{r}
#| label: mean-sale-price-for-each-category 
#| echo: true
#| message: false

# For nice looking tables!
library(gt)

companies %>% 
  spread_all() %>% 
  # Manually extracting information from acquisitions
  enter_object(acquisitions) %>% 
  gather_array() %>% 
  spread_all() %>% 
  # Removing missing values from relevant variables
  drop_na(category_code, price_currency_code, price_amount) %>% 
  # Finding max sale price for different company categories
  group_by(category_code, price_currency_code) %>%
  summarize(
    max_price = max(price_amount), 
    .groups = "drop"
            ) %>% 
  mutate(
    # Convert to dollar format
    max_price = scales::dollar(max_price), 
    # Remove _ from company sector
    category_code = str_replace_all(category_code, 
                                    pattern = "_",
                                    replacement = " "), 
    # Capitalize the first letter of each word
    category_code = str_to_title(category_code)
    ) %>% 
  gt() %>% 
  cols_label(category_code = md("**Business Sector**"), 
                 price_currency_code = md("**Currency of Sale**"), 
                 max_price = md("**Max Company Sale Price**")
                 )
```

**Step 5** -- Scaling to multiple columns

This is a powerful process, but you might be able to see how we don't
necessarily want to do this process repeatedly for every value `spread_all()`
was not able to parse.

::: callout-warning
# Do you see a way you could use iteration to complete this task?
:::

```{r}
#| label: attempting-to-iterate
#| eval: false

values_to_map <- companies_values %>% 
  filter(
    !name %in% colnames(companies_tidy)
      ) %>% 
  filter(name != "_id") %>% 
  pull(name) %>% 
  as.list()

manual_spread <- function(value){
  companies %>% 
  enter_object({{value}}) %>% 
  gather_array() %>% 
  spread_all()

}

manual_bind <- function(x){
  companies %>% 
    spread_all() %>% 
}

thing1 <- map(.x = values_to_map[c(1, 2)], 
              .f = manual_spread
              )

reduce(thing1, bind_cols)
```

<!-- ## Extra Resources -->

<!-- -   The GitHub page for [`rvest`](https://github.com/tidyverse/rvest).  -->
<!--     * Be sure to check out the vignettes and other documentation! -->
<!-- -   In-depth (for the scope of our course) [crash course in JSON](https://www.youtube.com/watch?v=wI1CWzNtE-M) -->
