---
title: "Team Season Dashboard"
format: dashboard
server: shiny
---

```{r}
#| context: setup
library(shiny)
library(tidyverse)
library(plotly)
library(leaflet)
library(gganimate)
library(gifski)

teams <- read_csv("https://www.dropbox.com/scl/fi/k02mwgl05fkqnuub0nmi0/nhl_team_data.csv?rlkey=k1r4n50bejrvm38dxna9y1lbw&st=xi3591j8&dl=1", 
                  show_col_types = FALSE)

metric_options <- c(
    "Points" = "P",
    "Wins" = "W",
    "Goals For" = "GF",
    "Goals Against" = "GA",
    "Goal Differential" = "goal_diff",
    "Win Percentage" = "win_pct"
  )

total_seasons <- n_distinct(teams$year)

total_teams <- n_distinct(teams$Team)

most_championships <- teams |>
  filter(playoff_result == "Won Stanley Cup") |>
  count(Team, sort = TRUE) |>
  slice_max(n, n = 1, with_ties = FALSE) |> 
  mutate(Team = word(Team)) |> 
  pull(Team)
```

# Team Trends

## Sidebar {.sidebar}

```{r}
selectInput(
  "team",
  "Team",
  choices = sort(unique(teams$Team)),
  selected = "San Jose Sharks"
)
```

```{r}
sliderInput(
  "year_range",
  "Years",
  min = min(teams$year, na.rm = TRUE),
  max = max(teams$year, na.rm = TRUE),
  value = range(teams$year, na.rm = TRUE),
  sep = ""
)
```

```{r}
selectInput(
  "metric",
  "Metric",
  choices = metric_options
)
```

## 

###  {height="30%"}

```{r}
#| content: valuebox
#| title: "Total Seasons in Data"

list(icon = "calendar-range", 
     color = "#2E86AB", 
     value = total_seasons)

```

```{r}
#| content: valuebox
#| title: "Total Teams in Data"

list(icon = "people-fill",
     color = "#4A4A4A",
     value = total_teams)

```

```{r}
#| content: valuebox
#| title: "Team with Most Championships"

list(icon = "trophy-fill",
     color = "#D4AF37",
     value = most_championships )
```

###  {.tabset}

```{r}
#| title: "Team Trends (Static)"
plotlyOutput("trend_plot")
```

```{r}
#| title: "Team Trends (Animated)"
imageOutput("trend_plot_anim",
            height = "50%",
            width = "50%")
```

```{r}
#| context: server
filtered <- reactive({
  teams |>
    filter(
      Team == input$team,
      year >= input$year_range[1],
      year <= input$year_range[2]
    )
})
```

```{r}
#| context: server
output$trend_plot <- renderPlotly({
  dat <- filtered() 
  team_color <- dat$Color[1]
  metric_name <- names(metric_options)[metric_options == input$metric]
  
  p <- dat |>
    ggplot(aes(x = year, 
               y = .data[[input$metric]],
               group = 1,
               text = playoff_result)) +
    geom_line(color = team_color) +
    geom_point(color = team_color) +
    labs(
      x = "",
      y = "",
      title = paste(input$team, metric_name , "by Season"),
    ) +
    theme_classic() 

  ggplotly(p, tooltip = "text")
})
```

```{r}
#| context: server

output$trend_plot_anim <- renderImage({
  
  dat <- filtered() 
  team_color <- dat$Color[1]
  metric_name <- names(metric_options)[metric_options == input$metric]
  
  p <- dat |>
    ggplot(aes(x = year, 
               y = .data[[input$metric]],
               group = 1,
               text = playoff_result)) +
    geom_line(color = team_color) +
    geom_point(color = team_color) +
    labs(
      x = "",
      y = "",
      title = paste(input$team, metric_name , "by Season"),
    ) +
    theme_classic() +
    transition_reveal(year)
  
    # Save as GIF
    tmpFile <- tempfile(fileext = ".gif")
    anim_save(tmpFile, 
              animate(p, 
                      renderer = gifski_renderer(), 
                      nframes = 20,
                      fps = 2,
                      res = 300,
                      height = 2,
                      width = 4,
                      units = "in"))
    
    # Return list with src
    list(src = tmpFile, contentType = "image/gif")
  }, deleteFile = TRUE)

```

# Team Locations

## Sidebar {.sidebar}

```{r}
selectInput(
  "season",
  "Season",
  choices = unique(teams$year)
)
```

## 

###  {height="25%"}

::: {.valuebox title="The NHL: Over a Century of Expansion"}
Founded in 1917 with just four Canadian teams, the NHL shrank to the famous Original Six by 1942---Boston, Chicago, Detroit, Montreal, New York, and Toronto. These teams comprised a stable core that lasted until 1967. That year's expansion doubled the league overnight, and successive waves through the 1970s and 1990s grew it to today's 32 teams spanning the US and Canada. Relocation has also reshaped the map you see below. In 1995 Quebec became Colorado, then in 1997 Hartford became Carolina, and finally in 2011 Atlanta became Winnipeg. Each of these changes reflect shifting markets and the sport's evolving geography.
:::

### 

```{r}
leafletOutput("team_map")
```

```{r}
#| context: server
output$team_map <- renderLeaflet({
    map_data <- teams |>
      filter(
        year == input$season
      ) |>
      distinct(Team, Long, Lat, Logo) |>
      drop_na(Lat, Long)

  leaflet(map_data) |>
    addProviderTiles(providers$CartoDB.Positron) |>
    addMarkers(
      lng = ~Long,
      lat = ~Lat,
      icon = ~icons(
        iconUrl = Logo,
        iconWidth = 60,
        iconHeight = 60
      ),
      label = ~Team
    )
})
```
