---
title: "Unit 3 Exam - Practice"
format: html
execute:
  eval: false
---

A programmer is making a package called `baRca` to help them analyze information about songs from their favorite soccer team, *FC Barcelona*.

The first four functions they create in the package are:

- `roster()` which takes a year, and returns a vector of names of all the players on the team in that year.

- `get_player_stats()`, which takes the name of a player as input and returns a tibble of statistics for that player (goals, assists, and minutes played) for each season they played for Barcelona.

- `team_results()`, which takes a year as input and returns a tibble of all the team's games in that year and the score.

- `best_player()`, which takes a year as input and returns the name of the player with the best points per minute (`(goals + assists)/minutes_played`) in that year.

### 1. Workflow

For each package creation step below, describe what it does and why it is necessary.

1.  `usethis::create_package("baRca")`.
2.  Edit the DESCRIPTION file
3.  `usethis::use_r("roster")`.
4.  Edit the \`roster.R\`\` file.
5.  Click `Code > Insert roxygen skeleton` in RStudio.
6.  Edit the `roster.R` file.
7.  `usethis::use_package("readr")`.
8.  `usethis::use_test("roster")`.
9.  Edit the `test-roster.R` file.
10. `Ctrl/Cmd-Shift-D` or `devtools::document()` or click the `Document` button in RStudio.
11. `Ctrl/Cmd-Shift-B` or `devtools::build()` or click the `Build` button in RStudio.
12. `Ctrl/Cmd-Shift-T` or `devtools::test()` or click the `Test` button in RStudio.

::: callout-note
# Solution

1.  Creates the "baRca" folder, the DESCRIPTION and NAMESPACE files, and the R/folder.
2.  Adds credit for authorship and descriptions of the package.
3.  Creates the `roster.R` file.
4.  The actual function code is written.
5.  Puts some placeholder roxygen-style comments in `roster.R`.
6.  Documents the inputs, outputs, and dependencies.
7.  Makes sure the `baRca` function installs `readr` too. Specifically, this adds a line in the DESCRIPTION that says `baRca` depends on `readr`.
8.  Creates the `testthat` folder setup and unit test file `test-roster.R`.
9.  Write some unit tests!
10. Uses the roxygen comments to automatically write `roster.md` in the `man/` folder.
11. Builds and installs the package.
12. Runs all the `testthat` unit tests.
:::

### 2. Unit testing

For the function `get_player_stats()`, describe:

1.  Two unit tests you would write using `expect_equal()`.

::: callout-note
# Solution

*Some possibilities include...*

```{r}
dat <- get_player_stats("Lionel Messi")

expect_equal(names(dat), c("player_name", "season", "goals", "assists", "minutes_played"))

expect_equal(class(dat), "data.frame")

# using information you'd have to look up, you could also do:

expect_equal(nrow(dat), 17)

goals_in_2017 <- dat |> 
  filter(year == 2017) |> 
  pull(goals)

expect_equal(goals_in_2017, 50)
```
:::

2.  Two unit tests you would write using `expect_error()`.

::: callout-note
# Solution

*This answer should test two different kinds of bad input, such as:*

```{r}
# no input
expect_error(get_player_stats())

# wrong input type
expect_error(get_player_stats(2017))

# wrong input structure
expect_error(get_player_stats(c("Lionel Messi", "Robert Lewandowski")))

# an input that is correct structure, but not correct for the data
expect_error(get_player_stats("Justin Bieber"))
```
:::

### 3. Data Management

The programmer is designing their `team_results()` function and has discovered that all the results of FC Barcelona's games can be found in the pages of [espn.com](espn.com). They are debating how they should use to access these data for their `team_results()` function.

Describe three *different* ways the programmer could design their package, their data storage, and/or their `team_results()` function to load the data from a particular year.

State which option (of the three) you would recommend they use and why.

::: callout-note
# Solution

1.  Write the function to **webscrape** from espn.com.

2.  Collect the data ahead of time from espn.com, and save it on GitHub or Dropbox. Then, write the function to read the `.csv` from a URL.

3.  Collect the data ahead of time from espn.com, and save it in the `inst/exdata` folder of the package. Then, in the function, use `system.file` and `read_csv()` to read the data from the package file.

The third option is probably best here, because it guarantees the data that is needed will always be available in the package.
:::

### 4. Documentation

The programmer's source code for the `get_player_stats()` function is below:

```{r}
#| eval: false


best_player <- function(year) {
  
  all_players <- roster(year) 
  
  all_stats <- map(.x = all_players, .f = get_player_stats) |> 
    bind_rows()
  
  all_stats |> 
    filter(season == year) |>
    mutate(
      ppm = (goals + assists)/minutes_played
    ) |>
    slice_max(ppm, with_ties = FALSE) |>
    pull(player_name)
  
}
```

Write a complete `roxygen2` style documentation for this function.

::: callout-note
# Solution

```{r}
#' Find the best player (in points per minute) from a given year
#' 
#' @param year A numeric value for the desired year
#' 
#' @return A single string with the name of the top player
#' 
#' @importFrom dplyr filter mutate slice_max pull bind_rows
#' @importFrom purrr map
#' 
#' @examples
#' 
#' best_player(2017)
#' 
#' @export
```
:::

### 5. Speed

Assume all functions are working as expected and are properly documented.

The programmer now uses their package to find the best player in each of the last 75 years, by running:

```{r}
map(.x = 1950:2025, .f = best_player)
```

Unfortunately, the programmer finds this code to be a bit slow.

Describe, in words or by sketching approximate code, three ways you would modify this code **or** one of the functions it uses to speed it up.

::: callout-note
# Solution

*Some possibilities include...*

- In the code above, change `map` to `furrr::future_map()`.

- In the `best_player()` function, change `map` to `furrr::future_map()`.

- Store the datasets of player statistics as `.parquet` files rather than `.csv`. Then, in the `get_player_stats()` function, use `arrow::read_parquet()` instead of `dplyr::read_csv()`.

- Replace the data wrangling code in the `best_player()` function with `data.table` syntax:

```{r}
all_stats <- data.table(all_stats)

all_stats[season == year, ppm := (goals + assists) / minutes_played]

all_stats[which.max(ppm), player_name]
```
:::

### 6. Package Design

Suggest two more functions for this package. For each, provide:

- The name and a brief description of what the function does.

> Example: "`roster` gets FC Barcelona roster for a given year."

- The argument(s) it expects as input and what it returns as output.

> Example: "Input is the year as a number, output is a vector of strings containing player names."

- An example of how and why someone might use the function.

> Example: "Someone might use this function to find out which players played the most seasons with the team."

::: callout-note
# Solution

*Answers will vary here, but we are looking for...*

1.  Functions that are feasible with the described package.

2.  A reasonable descriptive function name.

3.  A convincing use case to motivate the need for the function.

4.  Clearly defined object types and structures for the input and output.
:::
