---
title: "Unit 3 Exam - Practice"
format: html
execute:
  eval: false
---

A programmer is making a package called `baRca` to help them analyze information about songs from their favorite soccer team, *FC Barcelona*.

The first four functions they create in the package are:

- `roster()` which takes a year, and returns a vector of names of all the players on the team in that year.

- `get_player_stats()`, which takes the name of a player as input and returns a tibble of statistics for that player (goals, assists, and minutes played) for each year they played for Barcelona.

- `team_results()`, which takes a year as input and returns a tibble of all the team's games in that year and the score.

- `best_player()`, which takes a year as input and returns the name of the player with the best points per minute (`(goals + assists)/minutes_played`) in that year.

### 1. Workflow

For each package creation step below, describe what it does and why it is necessary.

1.  `usethis::create_package("baRca")`.
2.  Edit the DESCRIPTION file
3.  `usethis::use_r("roster")`.
4.  Edit the `roster.R` file.
5.  Click `Code > Insert roxygen skeleton` in RStudio.
6.  Edit the `roster.R` file.
7.  `usethis::use_package("readr")`.
8.  `usethis::use_test("roster")`.
9.  Edit the `test-roster.R` file.
10. `Ctrl/Cmd-Shift-D` or `devtools::document()` or click the `Document` button in RStudio.
11. `Ctrl/Cmd-Shift-B` or `devtools::build()` or click the `Build` button in RStudio.
12. `Ctrl/Cmd-Shift-T` or `devtools::test()` or click the `Test` button in RStudio.

### 2. Unit testing

For the function `get_player_stats()`, describe:

1.  Two unit tests you would write using `expect_equal()`.

2.  Two unit tests you would write using `expect_error()`.

### 3. Data Management

The programmer is designing their `team_results()` function and has discovered that all the results of FC Barcelona's games can be found in the pages of [espn.com](espn.com). They are debating how they should use to access these data for their `team_results()` function.

Describe three *different* ways the programmer could design their package, their data storage, and/or their `team_results()` function to load the data from a particular year.

State which option (of the three) you would recommend they use and why.

### 4. Documentation

The programmer's source code for the `get_player_stats()` function is below:

```{r}
#| eval: false

best_player <- function(year) {
  
  all_players <- roster(year) 
  
  all_stats <- map(.x = all_players, .f = get_player_stats) |> 
    bind_rows()
  
  all_stats |> 
    filter(year == year) |>
    mutate(
      ppm = (goals + assists)/minutes_played
    ) |>
    slice_max(ppm) |>
    pull(player_name)
  
}
```

Write a complete `roxygen2` style documentation for this function.

### 5. Speed

Assume all functions are working as expected and are properly documented.

The programmer now uses their package to find the best player in each of the last 75 years, by running:

```{r}
map(.x = 1950:2025, .f = best_player)
```

Unfortunately, the programmer finds this code to be a bit slow.

Describe, in words or by sketching approximate code, three ways you would modify this code **or** one of the functions it uses to speed it up.

### 6. Package Design

Suggest two more functions for this package. For each, provide:

- A brief description of what the function does.

> Example: "Gets FC Barcelona roster for a given year."

- The argument(s) it expects as input and what it returns as output.

> Example: "Input is the year as a number, output is a vector of strings containing player names."

- An example of how and why someone might use the function.

> Example: "Someone might use this function to find out which players played the most seasons with the team."

::: callout-important
# Solutions

You can access the practice exam "solutions" here: [Unit 3 Practice Exam Solutions](unit-3-practice-exam-solutions.qmd)
:::
