---
title: "Practice Activity Week 2"
format: html
editor: source
execute: 
  error: true
  eval: false
resources: 
  - student/pa-2.qmd
---

## Accessing the Practice Activity

Download the template Practice Activity Quarto file here: <a href="student/pa-2.qmd" download>`pa-2.qmd`</a>

::: callout-important
Be sure to save the file inside your Week 2 folder of your STAT 431 (or 541) folder!
:::

## Writing Functions

1.  Write a function called `summ_stats()` that takes a vector `x` as input and
produces the following output:

-   for **numeric variables**: returns the mean, median, standard deviation, and
IQR as a dataframe
-   for **categorical variables**: returns the number of levels (categories) of
the variable as a dataframe

*Hint:* You can use `tibble()` to create the data frame. For example, 
`tibble(a = 1:2, b = 2:3)` creates a data frame with variables `a` and `b`.

2. Confirm that your function works on each type of variable by running the 
code below. 

```{r}
summ_stats(diamonds$carat)

summ_stats(diamonds$color)
```

## Iterating Functions

3.  Use `map()` to apply your `summ_stats()` function to every column in the
`diamonds` dataset. 

*Hint: Look up the `bind_rows()` documentation from dplyr to combine summary
statistics for all the variables into one data frame. The `.id`
argument will be especially helpful in adding the variable names!*

## Data Transformation

4.  Let's make the output more intuitive, with the variables on the columns and
the summary statistics on the rows.

*Hint: You will need to do a double pivot (`pivot_longer()` then
`pivot_wider()`) to achieve this result!*

## Using Helper Functions (Stat 541 only)

5.  Now that we have the output we want, let's use our code to write a
`summ_df()` function that takes a data frame as an input and outputs a table of
summary statistics for every variable in the data frame.

*Hint: The body of your function should contain __all__ the code from Question 3.*

6.  Demonstrate that your function works using the `mpg` dataset.

```{r}
#| label: try-df-function

```
