---
title: "Practice Activity Week 3"
format: html
editor: source
execute: 
  eval: false
resources: 
  - student/pa-3.qmd
---

## Accessing the Practice Activity

Download the template Practice Activity Quarto file here: <a href="student/pa-3.qmd" download>`pa-3.qmd`</a>

::: callout-important
Be sure to save the file inside your Week 3 folder of your STAT 431 (or 541) folder!
:::

```{r}
#| label: packages-setup
#| message: false
#| warning: false

library(httr)
library(tidyjson)
library(jsonlite)
library(tidyverse)
```

# Open-Notify API

The Open-Notify APIs provides information on the International Space Station.
Specifically, the API provides (1) the names of the astronauts currently 
aboard the ISS, (2) the current location of the ISS, and (3) overhead pass
predictions for the ISS. 

For this activity, we are going to find the predicted time the ISS will pass
by 10 major cities across the world and sort the pass times to see the path 
of the ISS. 

| City        | Country        | Latitude | Longitude |
| ----------- | -------------- | -------- | --------- |
| New York    | USA            | 40.7128  | -74.0060  |
| Mexico City | Mexico         | 19.4326  | -99.1332  |
| São Paulo   | Brazil         | -23.5505 | -46.6333  |
| London      | United Kingdom | 51.5074  | -0.1278   |
| Cairo       | Egypt          | 30.0444  | 31.2357   |
| Nairobi     | Kenya          | -1.2921  | 36.8219   |
| Mumbai      | India          | 19.0760  | 72.8777   |
| Bangkok     | Thailand       | 13.7563  | 100.5018  |
| Tokyo       | Japan          | 35.6762  | 139.6503  |
| Sydney      | Australia      | -33.8688 | 151.2093  |

## Getting the Pass Times for *One* City

Inspecting the [satellite pass API homepage](https://g7vrd.co.uk/public-satellite-pass-rest-api), we can obtain the
JSON data with the pass information for a given city using an address like the 
following:
<https://api.g7vrd.co.uk/v1/satellite-passes/25544/51.45/-2.5833.json?hours=18>

In this web address, 

- `25544` stands for the NORAD Catalog Number for the ISS. So, this number will
**never** change for different cities. 
- `51.45` is the latitude for the given location
- `-2.5833` is the longitude for the given location
- `?hours=18` tells the API how many hours you want to consider for the passes

Let's give this a go! 

1. Using the `base_url`, `lat` and `long` objects, obtain the ISS pass times for
New York City. Save this object as `nyc`. 
*Hint 1: The `glue()` function from the __glue__ package or the `str_c()` function from __stringr__ will be helpful here!*
*Hint 2: What function do you use to get information from an API?*

```{r}
#| label: nyc-info

base_url <- "https://api.g7vrd.co.uk/v1/satellite-passes/25544/"

lat  <- 40.7128
long <- -74.0060
```

2. Now that we have the JSON file for NYC, let's parse it into a tidy dataframe.
Using functions from **jsonlite** or **tidyjson** convert the JSON file to a 
dataframe. 
*Hint: You might find it helpful to inspect the raw JSON file [here](https://api.g7vrd.co.uk/v1/satellite-passes/25544/40.7128/-74.006.json?hours=18).*

```{r}
#| label: extracting-nyc-passtime-info

```

3. The `start` column for the passtimes is what we are interested in, but
currently this is formatted as a string (e.g., `"2026-02-26T05:23:21.257Z"`). 
Parse this column into a data-time object using the `ymd_hms()` function from
**lubridate**. 
*Hint: The pass times for the ISS are reported in UTC (Coordinated Universal Time).* 

```{r}
#| label: parse-date-nyc

```

4. Our final step is to isolate the pass time for NYC. Use the `pull()` function
to extract the parsed data-time column you made in the last question. 
**Can you collapse _all_ the steps into _one_ pipeline?**

```{r}
#| label: pull-date-nyc

```

## Getting the Pass Times for *Every* City

Now that we have worked out each step using *one* city, the next step is to 
scale this process to **multiple** cities. We've created a tibble with ten 
major cities across the world. Your goal is to acquire the pass time for each
city.

```{r}
#| label: cities-to-map

cities <- tibble(
  city = c(
    "New York", "Mexico City", "Sao Paulo", "London",
    "Cairo", "Nairobi", "Mumbai", "Bangkok",
    "Tokyo", "Sydney"
  ),
  country = c(
    "USA", "Mexico", "Brazil", "United Kingdom",
    "Egypt", "Kenya", "India", "Thailand",
    "Japan", "Australia"
  ),
  latitude = c(
    40.7128, 19.4326, -23.5505, 51.5074,
    30.0444, -1.2921, 19.0760, 13.7563,
    35.6762, -33.8688
  ),
  longitude = c(
    -74.0060, -99.1332, -46.6333, -0.1278,
    31.2357, 36.8219, 72.8777, 100.5018,
    139.6503, 151.2093
  )
)
```

5. Our first step is to write a function that gets the pass time for *any* 
latitude and longitude coordinates. Using your code from Questions 1 though 4, 
create a `get_pass_time()` function which outputs the pass time for a given
location. 
*Hint 1: Your function should take __two__ arguments: latitude and longitude.*
*Hint 2: Think about how many objects you need to save!*

```{r}
#| label: pass-time-function

```

6. Using `map2()` iterate your function over the `latitude` and `longitude` 
columns of the `cities` dataframe. 
*Hint: There are three arguments to `map2()`, `.x`, `.y`, and `.f`. You will
need to specify each!*

```{r}
#| label: map-pass-times

```

Crap! It seems we got an error. Let's think through what might be happening. 

7. How does your function handle cities without pass times? What can you add to
the body of your function to handle this special case?

```{r}
#| label: pass-time-function-no-passes

```

8. Alright, let's try this again! Using `map2()` iterate your function over the
`latitude` and `longitude` columns of the `cities` dataframe. 

```{r}
#| label: map-pass-times-take-2

```

9. Your `map()` should have been successful, but you should notice something 
else happening. There are cities with **multiple** pass times! What can
you add to the body of your function to grab the *first* pass time for cities
with multiple pass times?

```{r}
#| label: pass-time-function-mult-passes

```

10. Alright, let's try this one last time! Using `map2()` iterate your function over the
`latitude` and `longitude` columns of the `cities` dataframe. 
**Save this object as `city_passes` so you don't get an error about exceeding 10 requests per minute.**

```{r}
#| label: map-pass-times-take-3

city_passes <- 
  
```

11. Add each city's pass times to the original `cities` dataframe. Arrange the
table based on the cities where the ISS will pass next. 

*Hint: The `city_passes` object is a list which you will need to unpack. The
`unnest()` will be helpful here!* 

```{r}
#| label: cities-with-passtimes


```
