---
title: "Practice Activity 9: Find the bugs"
editor: source
format: html
---

## Overview

The repository `Cal-Poly-Advanced-R/entomology` contains a package that is mostly complete and mostly correctly structured, that performs some simple analyses and summaries on a dataset concerning a few species of beetle.[^1]

[^1]: Lubischew, A.A. (1962) On the use of discriminant functions in taxonomy. Biometrics, 18, 455-477.

Unfortunately, this package contain **many bugs**, and your job is to track them down! It also is missing any `@example` documentation or `testthat` tests to help identify the issues.

To make things a little easier, there is exactly one problem in each of the nine different `.R` files (not counting `package.R`, which is just a formality). There is also one extra bonus problem in the `summary.R` file.

Run the following to download the package source code:

```{r}
#| eval: false
usethis::use_course("Cal-Poly-Advanced-R/entomology")
```


Then open `pa-9.qmd`.

## Step One: Understand goals

The primary goal of this package is to guess the species of an unknown beetle,
based on two measurements of its horns (or *aedeagus*): the width in microns 
and the angle in degrees.

The `entomology` package supplies a function called `beetle_classify_nearest()`,
which takes in a width and angle and returns the predicted species.  However,
this function doesn't work due to errors in its code and the functions it
depends on.

Begin by looking at the source code of `beetle_classify_nearest()`, which is in
the `classify.R` file.  In your own words, what do the steps of this function 
seem to be doing?

Then, notice that `beetle_classify_nearest()` depends on two new functions:
`beetle_species()` and `beetle_centroid()`.  What `.R` file are these in?  What
do they do?  What new functions to *they* depend on?  What do those do?  And so
on.

### Step Two: Find syntax errors

If our code is poorly formed, we even won't be able to run it to test it!

In your `entomology` package project, attempt to **build** the package.
*(Reminder: Ctrl/Cmd-Shift-B is a shortcut!)*

What error does this build encounter?  Fix it, then build your package!


### Step Three: Test basic functions.

* Run `beetle_data()`.  Does this seem to work as expected?  If not, fix it.

* Run `beetle_species()`.  Does this seem to work as expected?  If not, fix it.

Rebuild your package once you have fixed `data.R`.

*Remember, there is exactly __one__ bug in each file, except `summary.R` which
has two*

### Step Four: Test validation functions

Notice that the package provides two helper functions to make sure that users
input appropriate values to the main functions. These are in `validation.R`.

Try to run the following unit test:

```{r}
#| eval: false
validate_species("Con")
```

You'll find that the `validate_species` function is not found, even if your package is fully build and loaded.  This is because it is an **unexported** function - it was made to help the developer simplify their design, not so that users could access it and use it.  Notice that the `roxygen2` documentation comments above these functions do not include an `@export` line, and that these functions do not show up in the `NAMESPACE` file.

To access and run an unexported function, use triple colons `:::`, e.g.

```{r}
#| eval: false
entomology:::validate_species("Con")
```

Run some other unit tests on these functions until you discover the hidden bug.  

*Hint: Make sure to try every possible input that the function claims it can accept.*

### Step Five: Test small functions

Now, write your own unit tests for the basic calculation and data wrangling 
functions in `counts.R`, `correlation.R`, `centroid.R`, `angle.R`, and `filter.R`. 

These unit tests should include:

* The simplest possible function calls that you expect to *work correctly*.  

* A more complicated function call that you still expect to *work correctly*.

* One or more function calls that you expect to *cause an error*.  

Make sure to state what you *expect* the output to be **before you run the function**; 
and be sure to consider the *object types and structures* you expect, not just the
values.

Continue developing and running unit tests until you find the bug in each of
these files.  Once you have fixed all of them, re-build and re-load your package,
and then try the tests again to make sure they behave as desired.


### Step Six: Test complicated functions.

Now that you've fixed problems in the helper functions, we can revisit the main
ones.

Create unit tests as above for the functions in `classify.R` and `summary.R`.  
Use debugging techniques like `debugonce()` or `browse()` to find where problems are happening.

*Hint: While there is only one bug directly in this file, there is also a problem in
another file that causes this code to break, that you should track down via
your debugging.*


### Test Coverage (Stat 541 Only)

All your unit tests should be written using function from the `testthat`
package. You **do not** need to include them in the `\tests` folder. 

<!-- Run `covr::package_coverage(".")`.  You should have 100% coverage! -->

### Hints and Advice

-   Most of the errors in this package are *not* computational; instead, they are related to syntax problems, inconsistent object types and structures, etc.

-   The bugs are all in the code itself, not in the descriptions or documentation of the functions; you can trust that the documentation describes what the functions *should* do.

-   Don't overlook the power of "rubber duck debugging" - describe what the function does, step by step, to a partner and see if you notice anything amiss as you say it out loud.

-   Debugging is hard! If you find yourself stumped and frustrated, take a break and come back later with fresh eyes. And don't forget to work together with you peers - don't just split the list in half, discuss *each* bug together to track them down faster.
