---
title: "Practice Activity Week 5"
format: html
embed-resources: true
editor: source
---

```{r}
#| include: false
#| label: packages

library(tidyverse)
library(tigris)
```

# Data

`secret_map.csv` contains a secret code for each county in the U.S. In this 
Practice Activity, you will figure out the code and then create a map in
`ggplot` showing what you found.

```{r}
#| label: read-in-data

```

# Simple Map of US Counties

Let's start by just creating a map of all of the counties in the continuous US. 

1. Use the `counties()` function from the **tigris** package to load a dataframe
of US **counties** with their polygon files. Save this dataframe into an object
named `county_map`.

```{r}
#| label: question-1
#| message: false


```

2. Using `ggplot()`, generate a map of the US counties. Have the fill color for
each county be white and the outline color be a dark grey. Use cartesian
coordinates.
*Hint: `geom_sf()` will be helpful here!*

```{r}
#| label: question-2


```

3. Remove all axes labels and gridlines. 
*Hint: There is a built-in theme that can do this!*

4. While cartesian coordinates are typical, they are actually somewhat visually
misleading since the globe is a sphere. A recommended system for showing the US
is the "Albers Equal Area Conic Projection". Copy and paste your code from 
Questions 2 and 3 and change the coordinate system to this projection (or any
conic projection). 
*Hint: Check out the documentation for `coord_sf()` for other projection systems.*

```{r}
#| label: question-4
  
```


Great! Now you should have a nice simple map of all of the US counties. Let's 
add the information from the `secret_map.csv` data.

# Color that Map

To color the map we will need to join the `county_map` dataframe with the 
`secret_map` dataframe. We will break this up into a few steps since there are
some adjustments we need to make to each dataset before we join them 
together!

5. There are a couple of typos in the `secret_code` column of the `secret_map`
data. Let's fix these errors!
  a. Any "g" should be replaced with "F" (case matters!)
  b. The secret code should start with the hashtag character (#)
  
```{r}
#| label: question-5

```

6. Currently, the `county_map` dataframe doesn't have the names of the states, 
only their state code. Use the `fips_codes` object from the **tigris** package
to add the state names into the `county_map` dataframe. 
*Hint 1: You may want to join a dataframe smaller than the original `fips_codes`!*
*Hint 2: Pay attention to the column names between the two datasets!*

```{r}
#| label: question-6

```

7. Add the `secret_code` information to the `county_map` dataframe. 
*Hint: Pay attention to the column names between the two datasets!*

```{r}
#| label: question-7

```

8. Remove the counties with missing values in the `secret_code` column. 

```{r}
#| label: question-8

```

9. It turns out that the secret code gives you the colors that should be used in
your final map. Save a **vector** called `map_colors` with the unique
values of the `secret_code` column. *Hint: There should be 3 colors.*

```{r}
#| label: question-9

```

10. Now map it! Fill the counties as indicated in the secret map data with the
appropriate color (as given by `secret_code`) in your US map. 
*Hint 1: Remember the `secret_code` column gives both the variable to fill by in the plot AND the color to use.*
*Hint 2: You will want to use the `map_colors` vector to use the right colors.*

```{r}
#| label: question-10

```

# Canvas Submission: 

> Submit a screenshot of your final plot. 
> What flag is shown?