---
title: "Project Checkpoint 3"
execute: 
  eval: false
---

```{r}
#| label: packages-setup
#| message: false
#| warning: false

library(httr)
library(tidyjson)
library(jsonlite)
library(tidyverse)
library(glue)
```

## Setup

```{r}
#| include: false
#| echo: true

tuesdata <- tidytuesdayR::tt_load('2024-08-06')

olympics <- tuesdata$olympics
```

For simplicity here, I will study only the top 10 countries in terms of female medals. This code is from last week's project step.

```{r}
# Function to count medals but not overcount team sports

num_female_medals <- function(country) {
  
  female_medals <- olympics |>
    filter(team == country,
           sex == "F") |>
    distinct(games, event, medal) |>
    pull(medal)
  
  
  sum(!is.na(female_medals), na.rm = TRUE)
  
}
```

```{r}
top_countries <- olympics |>
  distinct(team) |>
  mutate(
    n_f_medals = map_dbl(team, num_female_medals)
  ) |>
  slice_max(n_f_medals, n = 10) |>
  pull(team)

top_countries
```

## New data from API

*I'm choosing to use the API-Ninjas API for country populations. This API requires an account and API key. In the code that you see below, I have replaced hidden the line that stores my API key, so that it is not publicly shared.*

```{r}
#| echo: false
my_key <- "1ekT3dWHWPDuwhNECaYVFxJqyiEDVZo6HJqPh9xF"
```

```{r}
get_population <- function(country, key = my_key)
  {
  
  url <- glue("https://api.api-ninjas.com/v1/population?country={country};X-Api-Key={key}")
  my_data <- safely(fromJSON)(url)
  
  if (is.null(my_data$result)) {
    return(NA)
  } else {
    return(my_data$result$historical_population)
  }
}

get_population("Japan")
get_population("United States")
```

Since "United States" has a space in it, it isn't being searched properly by the API URL. After some investigation on the API site and experimentation, I see that I can use country abbreviations instead.

```{r}
top_countries <- olympics |>
  distinct(team, noc) |>
  mutate(
    n_f_medals = map_dbl(team, num_female_medals)
  ) |>
  slice_max(n_f_medals, n = 10) |>
  pull(noc)

top_countries
```

```{r}
#| warning: false


data_list <- top_countries |>
  map(get_population)

# Check for ones that didn't work
data_list |> is.na() |> sum()
```

Now let's make this a tibble so we can join it to our original data:

```{r}
pop_df <- data_list |>
  set_names(top_countries) |>
  bind_rows(
    .id = "noc"
  ) 

pop_df |> head()
```

Now, we join the datasets:

```{r}
olymp_top <- olympics |>
  filter(noc %in% top_countries) |>
  left_join(pop_df)

olymp_top |> head()
```

Note: There are a lot of NAs because the population data doesn't go as far back in time as the olympic data!

Finally, we use our new information to produce an interesting summary:

```{r}
olymp_top |>
  summarize(
    n_gold = sum(medal == "Gold", na.rm = TRUE),
    population = population[1],
    .by = c(noc, sex, year)
  ) |>
  mutate(
    n_gold_pp = n_gold/population
  ) |>
  filter(sex == "F") |>
  group_by(noc) |>
  summarize(
    n_gold_pp = median(n_gold_pp, na.rm = TRUE)
  ) |>
  arrange(desc(n_gold_pp))
```

<!-- ## New data from webscraping -->

<!-- ```{r} -->

<!-- ``` -->

<!-- Now that you have explored your original dataset and produced data summaries addressing your primary research questions, it is time to integrate new country-level data. -->

<!-- ## By the end of the week, your updated report should contain: -->

<!-- -   **two** new country-level data sources -->

<!--     -   Stat 431: One of these must be acquired using APIs or webscraping -->

<!--     -   Stat 541: One of these must be acquired using APIs, the other using webscraping. -->

<!-- -   descriptions of each of the new datasets -->

<!-- -   a "meta" dataset containing **all** the acquired data (joined together) -->

<!-- -   **at least two** data summaries OR visualizations incorporating the additional country-level data -->

<!-- ## Stat 541 Only -->

<!-- -   your **webscraped** data must come from a multi-paged source, so that iteration is used to gather all the data -->

<!-- ::: callout-warning -->

<!-- **Warning!** Most APIs require you to make an account to access them, to prevent too much automated data collection. *Some* APIs also charge a fee to use - you should *not* use any of these for this class. -->

<!-- ::: -->

<!-- ## Helpful Links -->

<!-- Here are some sources for country-level data that you may use if you so choose. -->

<!-- *These are just to get you started. You are not obligated to use these, and we cannot guarantee that they will all be functional, free, or useful.* -->

<!-- ### Tabular and/or Scrapable Data -->

<!-- -   [World Happiness Report](https://www.worldhappiness.report/data-sharing/) -->

<!-- -   [NOAA Climate Data](https://www.ncei.noaa.gov/cdo-web/) -->

<!-- -   [BEA Employment Data](https://www.bea.gov/data/employment/employment-county-metro-and-other-areas) -->

<!-- -   [WorldBank Development Indicator Data](https://databank.worldbank.org/id/f1ddbc94) -->

<!-- -   [Wikipedia - Temperatures](https://en.wikipedia.org/wiki/List_of_countries_by_average_yearly_temperature) -->

<!-- -   [Wikipedia - Populations](https://en.wikipedia.org/wiki/List_of_countries_and_dependencies_by_population) -->

<!-- -   [Wikipedia - Many demographics](https://en.wikipedia.org/wiki/Demographics_of_the_world) -->

<!-- ### APIs -->

<!-- -   [Currency, flags, capitals](https://restcountries.com/) -->

<!-- -   [Demographics and financial data](https://api-ninjas.com/api/country) -->

<!-- -   [WorldBank Data API](https://documents.worldbank.org/en/publication/documents-reports/api) -->

<!-- ::: callout-note -->

<!-- Much of the data we have relied on for years - such as NOAA and BEA data listed above - are no longer being collected and made available, due to US Government cuts to these public statistics services. Other data sources that result from independent academic research at US institutions are rapidly dwindling as well, also due to cuts to this funding.\ -->

<!-- \ -->

<!-- Although these data sources may eventually be restored, or replaced with data from alternative agencies, we will never be able to go back and fill the gap for these lost years. -->

<!-- ::: -->
