---
title: "Practice Activity 6 Warm-up"
format: docx
---

```{r}
library(tidyverse)
```

# Option 1: Latitudes & Longitudes

## California Counties

Notice Alameda has **multiple** rows in these data! What do you think each 
unique `long` and `lat` correspond to?

## How would you expect this plot to look? Draw it out!

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: false

ggplot(data = ca_counties, 
       mapping = aes(long, lat)
       ) + 
  geom_point(size = 0.25)
```

## What did we need to add to the code to get the plot to look like a map?

```{r}
#| echo: false
#| eval: false

ggplot(data = ca_counties, 
       mapping = aes(long, lat)
       ) + 
  geom_point(size = 0.25) +
  coord_sf()
```

## Now suppose I used `geom_polygon()` instead.

How would you expect this plot to look? Draw it out!

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: false

ggplot(data = ca_counties, 
       mapping = aes(long, lat, group = group)
       ) +
  geom_polygon(fill = "white", colour = "grey50") + 
  coord_sf()

```

# Option 2: Shape Files

## The `tigris` Package

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: false

library(tigris)

ca_counties <- tigris::counties(state = "california")
```

## Using Shape Files

::: midi
```{r}
#| label: shapes-with-coord-sf
#| echo: true
#| eval: false
#| code-line-numbers: false

ggplot(data = ca_counties) +
  geom_sf(fill = "white", 
          color = "grey25", 
          linewidth = 0.1) +
  theme_void()
```
:::

## Why not use `leaflet`?

| When to use `ggplot` to make a map | When to use `leaflet` to make a map |
|------------------------------------|---------------------------------------|
|                                    |                                       |
|                                    |                                       |
|                                    |                                       |
